### 2020-12-23 manuscript.tex
Add manuscript tex file

\documentclass[11pt,final,fleqn]{article}\usepackage[]{graphicx}\usepackage[]{color}

\usepackage{alltt}

% basic packages
\usepackage[T1]{fontenc}
\usepackage[margin=1in] { geometry }
\usepackage{amssymb,amsmath, bm}
\usepackage{verbatim}
\usepackage[latin1]{inputenc}
%\usepackage[OT1]{fontenc}
\usepackage{setspace}
\usepackage{natbib}
\usepackage{enumitem}
\usepackage[hyphens,spaces,obeyspaces]{url}
\usepackage[font={bf}]{caption}
%\usepackage{pgfplots}
%\usepackage[font={bf}]{caption}
\usepackage{latexsym}
%\usepackage{euscript}
\usepackage{graphicx}
\usepackage{marvosym}
%\usepackage[varg]{txfonts}  Older version of ``g'' in math.
\usepackage{pdflscape}
\usepackage{algorithm}

% bibliography packages
\usepackage{natbib}
\bibpunct{(}{)}{;}{a}{}{,}
\bibliographystyle{apa}
\renewcommand{\bibname}{References}

% hyperref options
\usepackage{soul} % To highlight text (along with color)
\usepackage{color}
\usepackage{hyperref}
\usepackage{xcolor}
\hypersetup{
    colorlinks,
    linkcolor={blue!50!black},
    citecolor={blue!50!black},
    urlcolor={blue!80!black}
}
\hypersetup{pdfstartview={XYZ null null 1.00}} % PDF opens to 100% zoom
\newcommand*{\Appendixautorefname}{Appendix}
\renewcommand*{\sectionautorefname}{Section}
\renewcommand*{\subsectionautorefname}{Section}
\renewcommand*{\subsubsectionautorefname}{Section}
\newcommand{\subfigureautorefname}{\figureautorefname}
\newcommand{\aref}[1]{\hyperref[#1]{Appendix~\ref{#1}}}
\newcommand{\algorithmautorefname}{Algorithm}

% packages for tables
\usepackage{longtable}
\usepackage{booktabs, threeparttable}
\usepackage{threeparttablex}
\usepackage{tabularx}
% dcolumn package
\usepackage{dcolumn}
\newcolumntype{.}{D{.}{.}{-1}}
\newcolumntype{d}[1]{D{.}{.}{#1}}
\captionsetup{belowskip=10pt,aboveskip=-5pt}
\usepackage{multirow}
% rotating package
\usepackage[figuresright]{rotating}
\usepackage{pdflscape}
\usepackage{subcaption}
\usepackage{caption} 
\captionsetup[table]{skip=5pt}

% packages for figures
\usepackage{grffile}
\usepackage{afterpage}
\usepackage{float}
\usepackage[section]{placeins}
\usepackage[export]{adjustbox}

% pacakges for code
\usepackage{listings}

% theorem package
\usepackage{theorem}
\theoremstyle{plain}
\theoremheaderfont{\scshape}
\newtheorem{theorem}{Theorem}
\newtheorem{assumption}{Assumption}
\newtheorem{lemma}{Lemma}
\newtheorem{proposition}{Proposition}
\newtheorem{remark}{Remark}
\newcommand{\qed}{\hfill \ensuremath{\Box}}
\newcommand\indep{\protect\mathpalette{\protect\independenT}{\perp}}
\DeclareMathOperator{\sgn}{sgn}
\DeclareMathOperator{\tr}{tr}
\DeclareMathOperator{\argmin}{arg\min}
\DeclareMathOperator{\argmax}{arg\max}
\def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}
\providecommand{\norm}[1]{\lVert#1\rVert}
\renewcommand\r{\right}
\renewcommand\l{\left}
\newcommand\E{\mathbb{E}}
\newcommand\dist{\buildrel\rm d\over\sim}
\newcommand\iid{\stackrel{\rm i.i.d.}{\sim}}
\newcommand\ind{\stackrel{\rm indep.}{\sim}}
\newcommand\cov{{\rm Cov}}
\newcommand\var{{\rm Var}}
\newcommand\SD{{\rm SD}}
\newcommand\bone{\mathbf{1}}
\newcommand\bzero{\mathbf{0}}
\DeclareMathOperator{\logit}{logit}
\DeclareMathOperator{\Cat}{Cat}
\DeclareMathOperator{\Multinomial}{Multinomial}

% file paths and definitions
\input{txtstats.txt}
\makeatletter
\newcommand*\ExpandableInput[1]{\@@input#1 }
\makeatother

% spacing 
\usepackage[compact]{titlesec}
\usepackage{etoolbox}
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setstretch{1.3}
\setlength{\abovecaptionskip}{15pt plus 3pt minus 2pt} % Space between caption and figure
\BeforeBeginEnvironment{equation}{\begin{singlespace}}
\AfterEndEnvironment{equation}{\end{singlespace}\noindent\ignorespaces}
\BeforeBeginEnvironment{align}{\begin{singlespace}}
\AfterEndEnvironment{align}{\end{singlespace}\noindent\ignorespaces}

% appendix settings
\usepackage[toc,page,header]{appendix}
\renewcommand{\appendixpagename}{\centering Appendices}
\usepackage{chngcntr}
\usepackage{lipsum}

% new commands
\newcommand\CPP{{C\texttt{++}}}
\newcommand\R{{\textsf{R}}}
\newcommand*\sameaff[1][\value{footnote}]{\footnotemark[#1]}
\newcommand\floor[1]{\lfloor#1\rfloor}
\newcommand\ceil[1]{\lceil#1\rceil}
\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\pkg}[1]{{\fontseries{b}\selectfont #1}}

% subsubsection
\titleclass{\subsubsubsection}{straight}[\subsection]

\newcounter{subsubsubsection}[subsubsection]
\renewcommand\thesubsubsubsection{\thesubsubsection.\arabic{subsubsubsection}}
\renewcommand\theparagraph{\thesubsubsubsection.\arabic{paragraph}} % optional; useful if paragraphs are to be numbered

\titleformat{\subsubsubsection}
  {\normalfont\normalsize\bfseries}{\thesubsubsubsection}{1em}{}
\titlespacing*{\subsubsubsection}
{0pt}{3.25ex plus 1ex minus .2ex}{1.5ex plus .2ex}

\makeatletter
\renewcommand\paragraph{\@startsection{paragraph}{5}{\z@}%
  {3.25ex \@plus1ex \@minus.2ex}%
  {-1em}%
  {\normalfont\normalsize\bfseries}}
\renewcommand\subparagraph{\@startsection{subparagraph}{6}{\parindent}%
  {3.25ex \@plus1ex \@minus .2ex}%
  {-1em}%
  {\normalfont\normalsize\bfseries}}
\def\toclevel@subsubsubsection{4}
\def\toclevel@paragraph{5}
\def\toclevel@paragraph{6}
\def\l@subsubsubsection{\@dottedtocline{4}{7em}{4em}}
\def\l@paragraph{\@dottedtocline{5}{10em}{5em}}
\def\l@subparagraph{\@dottedtocline{6}{14em}{6em}}
\makeatother

\setcounter{secnumdepth}{4}
\setcounter{tocdepth}{4}

% title
\title{Penalized regression for left-truncated and right-censored survival data}
\author{Author list}
\date{\today}

\providecommand{\keywords}[1]
{
  \small	
  \textbf{\textit{Keywords---}} #1
}
\begin{document}
\maketitle

\begin{abstract}
    High-dimensional data are common in the medical field as high-throughput screening, detailed electronic health record (EHR) information, and comprehensive genomics testing are now used to capture the patient profile. Statistical models that attempt to study the effects of many predictors on survival typically implement feature selection or penalized methods to mitigate the undesirable consequences of overfitting. In some cases survival data is also left truncated which can create an immortal time bias, but penalized survival methods that adjust for left truncation are unavailable in common software. To address these challenges, we implement a penalized Cox proportional hazards model for left-truncated and right-censored survival data and assess implications of left truncation adjustment on bias and inference. We use simulation studies and examples from high-dimensional real-world EHR and genomics data to highlight the pitfalls of failing to account for left truncation in survival modeling. The extension of survival modeling to left truncated data is available in the latest release of \pkg{glmnet}.
\end{abstract}

\keywords{left truncation, penalized regression, Lasso, high-dimensional data, survival analysis, Cox model}

\section{Introduction}
The recent development and accessibility of genomic (DNA sequencing, microarrays), proteomic (immuno-histology, tissue arrays), digital (wearables), and imaging (PET, fMRI) technologies have led to an increasing complexity and dimensionality of patient health information. The level of person-level and disease-specific detail captured by these technologies offers an exciting new opportunity to understand patient history and prognosis by combining these molecular data with macro-scale clinical data on patient characteristics, treatment, and outcomes. Specifically, these data may offer insights into risk factors affecting patient-related outcomes leading to interest in developing prognostic tools to improve clinical decision-making and patient care \citep{gui2005penalized, wishart2010predict, ow2016big, yousefi2017predicting}. Given the hundreds or even thousands of molecular, patho-physiological, and clinical parameters that may be readily available for each patient, developing an appropriate statistical model that is robust to overfitting is critical to ensure good model performance and generalizability. Survival analysis is of particular interest given that overall survival is often a primary outcome of interest.

In a high-dimensional setting, computationally efficient regularization and shrinkage techniques have been developed \citep{tibshirani1997lasso, friedman2010regularization, simon2011regularization} for the Cox model. Similar to ordinary least squares regression (OLS), these penalized regression methods estimate regression coefficients by minimizing the sum of squared errors, but place a constraint in the equation to minimize the sum of coefficients. This constraint (controlled by a parameter, $\lambda$) penalizes a high number of variables in the model and shrinks the coefficients of less influential features towards zero - acting, in cases when some coefficients are shrunk fully to zero, as automatic feature selection. Common penalized approaches include lasso \citep{tibshirani1996regression}, ridge \citep{hoerl1970ridgeA, hoerl1970ridgeB}, and elastic net \citep{zou2005regularization}, with extensions such as the adaptive lasso \citep{zou2006adaptive}.

A common challenge in survival data is that patients are often included in the data only after having been at risk (e.g. diseased) for some time so that patients with shorter survival times are unobserved. In other words, the data are left truncated because patients are only included if they survive beyond some initial milestone that marks entry into the observed data. An analysis of the observed patients is subject to an immortal time bias because observed patients cannot die prior to entering the study cohort \citep{levesque2010problem, giobbie2013challenges}. An example in medicine is a dataset describing patients who receive a specific diagnostic or genetic test; patients who die before having the chance to be tested are excluded from the data so their survival times do not contribute to overall survival estimates. Traditional survival models are inappropriate in this case and predicted survival probabilities are overestimated because of the exclusion of short survival times.

Left truncation is a well known feature of medical data and there are methods to adjust for it \citep{kalbfleisch2011statistical}. The Kaplan-Meier estimator of the survival function requires almost no modifications as it is only necessary to change the risk set so that individuals are only at risk of an event after study entry \citep{tsai1987note}. Similarly, with left truncation, the likelihood function in a parametric model can be adjusted so that it is conditional on survival to the entry time.

Here, we implement a penalized Cox proportional hazards for left-truncated and right-censored survival data and critically assess scenarios in which bias arises from and is exacerbated by left truncation. We apply the method to simulated data and a real-world use case from the Flatiron Health and Foundation Medicine Clinico-Genomic Database (CGDB). Our results highlight the importance of adjusting for left truncation for both inference and prediction. Furthermore, we show that use of inappropriate metrics for model evaluation can result in overly optimistic conclusions about model performance. For implementation of the proposed method, we introduce the left truncation extension to survival modeling in the R package \pkg{glmnet} v.X.X (CITE LATEST PACKAGE).

The remainder of this paper is organized as follows: In \autoref{sec:methods}, we introduce the penalized regression and left truncation framework and present the efficient method that adjusts for both. \autoref{sec:simulation} presents the simulation results and in \autoref{sec:rwd} the method is applied to a real-world dataset combining electronic health records and genomic testing. Finally, we discuss general takeaways and survival modeling recommendations in \autoref{sec:discussion}.

\section{Methods} \label{sec:methods}

\subsection{Data and notation} 
Consider a typical survival analysis framework in the absence of truncated survival times. For each patient $i$, we have data of the form $y_i$, $e_i$, $x_i$ where $y_i$ is the observed survival time, defined as the time between baseline and event, $e_i$ is the event or failure of interest (e.g. death), and $x_i = (x_{i,1}, x_{i,2} \ldots, x_{i,p})$ is a vector of covariates of length $p$ used in the regression model. The event or failure of interest $e_i$ is 1 if the event is observed or 0 if the event is right-censored. Suppose there are $m$ unique event times, and let $t_1 < t_2 < ... < t_m$ be the ordered, unique event times. In this framework, all subjects are assumed to be at risk of the event starting at a commonly-defined ``time zero'' (a milestone that serves as a reference point for survival such as date of diagnosis) and observed over the interval $[0,T]$ where $T$ is the maximum follow-up time. Let $j(i)$ be the index of the observation with an event at time $t_i$. A Cox proportional hazards model is commonly used to model the relationship between survival times $y_i$ and predictors $x_i$ assuming a semi-parametric hazard of the form

\begin{equation}
h_i(t) = h_0(t)e^{x_i^\intercal\beta}
\end{equation}

where $h_i(t)$ is the hazard for patient $i$ at time $t$, $h_0(t)$ is the baseline hazard shared across subjects, and $\beta$ is a length $p$ vector of predictor coefficients.

\citet{cox1972regression} proposed a partial likelihood for $\beta$ without involving the baseline hazard $h_0(t)$

\begin{equation} \label{eqn:cox1972}
\prod_{i=1}^{m} \frac{e^{x_{j(i)}^\intercal\beta}}{\sum_{j(i)\in R_i}  e^{x_{j(i)}^\intercal\beta}   } 
\end{equation}

where  $R_i = \{ j : y_j \geq t_i\}$ denotes the set of indices $j(i)$ who are ``at risk'' for failure at time $t_i$, called the risk set. Maximizing the partial likelihood solves for $\beta$ and allows for inferences on the relationship between survival time and the set of predictors.

\subsection{Left truncation}
In the absence of left truncation, the risk set $R_i = \{ j : y_j \geq t_i \}$ assumes that every individual is at risk of an event at time 0 and continues to be at risk until the event occurs or the individual is censored. In this sense, all individuals enter the risk set at the same time and leave only at death or censoring. However, when survival times are left truncated, individuals are truly at risk prior to entering the study (e.g. prior to being observed for follow-up). Take \autoref{fig:followup} as an example.

\begin{figure}[h]
\caption{Patient follow-up. Left-truncated, right-censored patient follow-up in a hypothetical study cohort. Patients become eligible to enter the study after diagnosis (closed circle) and are followed until death (open circle) or censoring (cross). However, patients who die before the start of the study (dashed blue line) are left truncated (in red), and only those who have survived until the study start (in black) are observed.}
\label{fig:followup}
\centering
\includegraphics[max size={\textwidth}{\textheight}]{figs/followup.pdf} 
\end{figure}

Survival is defined as the time from diagnosis until death, but the study does not begin until 2011 (dashed blue vertical line). Patients 2 and 5 (in red) are excluded from the analysis because of death and a right censoring event, respectively, prior to study eligibility. A naive analysis on this data would therefore be biased because it only contains survival time for ``super-survivors`` (those who have survived a long time) and the recently-diagnosed.

Fortunately, the partial likelihood can be easily modified to make $R_i$ the set of individuals who are at risk (have entered the study and are not censored or failed) at that time. Let individual $i$ have survival data of the form $(v_i, y_i, e_i, x_i)$, where individual $i$ enters the study at time $v_i$. The left-truncated risk set $R_{i\ast} = \{ j : y_j \geq t_i > v_j \}$. In other words, subjects are only considered at risk in the analysis after they have been observed to be at risk (e.g. exceeded their left truncation time). Here, event times $t_1 < t_2 < ... < t_m$ are ordered relative to the earliest chronological event time observed in the patient cohort, rather than assuming that all patients are at risk starting at the same referent time $0$.

The partial likelihood in \autoref{eqn:cox1972} can then be evaluated by substituting $R_{i\ast}$ to make inferences on $\beta$ in the case of left truncation.

\subsection{Penalized regression with left-truncated and right-censored data}
Penalized regression methods such as $l_1$ (lasso) \citep{tibshirani1996regression} and $l_2$ (ridge) \citep{tikhonov1963ridge} regression, elastic net regression \citep{zou2005regularization}, and SCAD (Smoothly Clipped Absolute Deviation) \citep{xie2009scad} have been developed to overcome the limitation of fitting a model to a large number of covariates. In a high dimensional feature space, where the number of predictors is larger than the sample size (i.e., $p$ > $n$), a model may exhibit low bias but suffer from poor prediction accuracy and limited generalizability, a consequence of overfitting the data. Penalized approaches, which add a regularization penalty to constrain the size of the $\beta$, reduce the complexity of the model and mitigate the negative effects of overfitting.

The extension of penalized regression methods to survival data have been extensively described and applied, for example in microarray gene expression data \citep{gui2005penalized}, transcriptomes \citep{wu2011penalized}, and injury scale codes \citep{mittal2013penalized}. Like linear regression, the performance and fit of a Cox proportional hazards model is known to deteriorate with large $p$, providing an advantage to penalized approaches in high-dimensional, time-to-event feature sets.

\citet{simon2011regularization} introduced a fast, pathwise algorithm to fit a regularized Cox proportional hazards model solving for

\begin{equation} \label{eqn:simon}
\hat{\beta} = argmax_{\beta}  \left[ \frac{2}{n} \left( \sum_{i=1}^m x_{j(i)}^\intercal\beta - log\left(\sum_{j(i)\in R_i}  e^{x_{j(i)}^\intercal\beta)}\right) \right)  - \lambda P_{\alpha}\beta  \right] 
\end{equation}

where $\beta$ is the vector of regression coefficients and is maximized over \autoref{eqn:simon} subject to a constraint defined by the penalty $\lambda \cdot P_{\alpha}\beta$. As described elsewhere \citep{simon2011regularization}, the penalty can take the form of the $l_1$ (lasso), $l_2$ (ridge), or elastic net penalty,  and the algorithm fits over a path of $\lambda$ values via cyclical coordinate descent. Instead of solving a least squares problem at each iteration, the algorithm solves a penalized reweighted least squares problem, in which observation weights $w$ are defined in the coordinate descent least squares minimization step. The optimal $\lambda$ can be identified through cross-validation. Tied survival times are handled via the Breslow approximation \citep{breslow1972} and details provided in \citep{simon2011regularization}.

Applying the standard penalized Cox implementation in the software \pkg{glmnet} to left-truncated survival data will result in biased estimates. However, the extension to adjust for left truncation is straightforward. Given survival data with left truncation and right censoring, we solve \autoref{eqn:simon} with a modification to the risk set $R_i$:

\begin{equation}
\hat{\beta} = argmax_{\beta}  \left[ \frac{2}{n} \left( \sum_{i=1}^m x_{j(i)}^\intercal\beta - log\left(\sum_{j(i)\in R_{i\ast}}  e^{x_{j(i)}^\intercal\beta)}\right) \right)  - \lambda P_{\alpha}\beta  \right] 
\end{equation}

where the left-truncated risk set $R_{i\ast} = \{ j : y_j \geq t_i > v_j \}$; that is, it considers subjects at risk only after their left truncation times. The method can take lasso, ridge, and elastic net penalties for $\lambda P_{\alpha}\beta$ defined elsewhere \citep{friedman2010regularization} and is a simple extension of default methods in \pkg{glmnet}. The efficient implementation of this method is non-trivial, as the usual updating of risk set components cannot be used when left truncation is present. This implementation is available in the release of \pkg{glmnet} v. X.X (CITE LATEST PACKAGE).

\subsection{Real-world use case}
We consider the Clinico-Genomic Database (CGDB) offered jointly by Flatiron Health and Foundation Medicine. The CGDB is an oncology research platform that combines real-world, patient-level clinical data and outcomes with patient-level genomic data to provide a comprehensive patient and tumor profile. Patients in the CGDB have received at least one Foundation Medicine tumor genomic test as well as have had their electronic health records captured by Flatiron Health. Together, these data sources represent thousands of potential predictors per patient, ranging from demographic, lab, and treatment information to the specific mutations detected on each biomarker included in a Foundation Medicine baitset.

We take a subset of $\nPatients$ stage IV, non-small cell lung cancer (NSCLC) patients in the CGDB as a use case for this study. The outcome of interest is survival time predicted from the date of Stage 4 diagnosis. Survival times are left-truncated because patients who die before receiving a Foundation Medicine test are logically not included in the cohort of Foundation Medicine test recipients.

As shown in \autoref{fig:left-trunc-time} the distribution of left truncation times (time from diagnosis to genomic testing) is right-tailed with approximately $\entryMoreOneYear$ of patients diagnosed more than one year prior to receiving the Foundation Medicine genomic test.  Moreover, the correlation between diagnosis year and left truncation time is strongly negative ({$\rho$} = $\corrDxYear$), suggesting the presence of left truncation survivor bias as patients with records in the CGDB diagnosed longer ago needed to survive more years before being tested.

\begin{figure}[h]
\caption{Distribution of left truncation time (days) in non-small cell lung cancer patients in the CGDB}
 \label{fig:left-trunc-time}
 \centering
\includegraphics[max size={\textwidth}{\textheight}]{figs/left_trunc_hist.pdf} 
\end{figure}

Next, we use these data as the basis for a simulation study and a real-world application.

\section{Simulation study} \label{sec:simulation}
Simulations were conducted to assess the performance of the method with left truncated survival data. To set notation, let $T_i$, $U_i$, and $V_i$ be latent survival, right censoring, and study entry times for patient $i$, respectively. The simulation proceeds by simulating $T_i$, $U_i$, and $V_i$ from suitable survival distributions and assumes that $T_i$, $U_i$, and $V_i$ are uncorrelated. The observed survival time is given by $Y_i = \rm{min}(T_i, U_i)$ and a patient is right-censored if $T_i > U_i$. Patients are left-truncated (i.e., hidden from the sample) if $V_i > Y_i$.

\subsection{Data generating process}
We generated $n$ patients and $p$ predictors to construct a $n \times p$ matrix, $X$. The first 11 predictors were from the CGDB and the remaining $\tilde{p}$ variables were simulated binary variables. The CGDB variables were the same $11$ variables used in the small $p$ model in \autoref{sec:rwd} and generated by randomly sampling rows of the original data with replacement. The matrix of binary variables, $\tilde{X}$ was simulated using a (latent) multivariate probit approach

\begin{equation}
\begin{pmatrix}
\tilde{x}_{i,1}^{*} \\
\vdots \\
\tilde{x}_{i,\tilde{p}}^{*}
\end{pmatrix}
\sim
N
\left(\begin{pmatrix}
\mu_{1} \\
\vdots \\
\mu_{\tilde{p}}
\end{pmatrix},
\Sigma
\right),
\end{equation}

with the $k$th binary variable $\tilde{x}_{i,k}=1$ if $\tilde{x}_{i,k}^{*} > 0$ and $0$ otherwise. Each diagonal element of $\Sigma$ equals 1 and non-diagonal elements determine the correlation between variables. The probability, $\pi_k$ that  variable $k$ was equal to $1$ was randomly sampled from a uniform distribution ranging from .2 to .8; that is, we sampled $\pi_k \sim U(0.2, 0.8)$ and set $\mu_k = \Phi^{-1}(\pi_k)$ where $\Phi(\cdot)$ is the cumulative density function of the normal distribution. $\Sigma$ was generated by filling a $\tilde{p} \times \tilde{p}$ matrix, $\tilde{\Sigma}$, with $N(0,1)$ random variables and setting  $\Sigma = S/D^TD$ with $S = \tilde{\Sigma}^T\tilde{\Sigma}$ and $D$ a vector containing the diagonal elements of $S$.

Latent survival time was simulated from a proportional hazards Weibull distribution with shape parameter $a$ and scaled parameter $m_i$ based on evidence from the CGDB as detailed in \autoref{subsec:sim-calibration}. Covariates predict latent survival time through their effect on the scale parameter,

\begin{align} 
m_i &= \rm{exp}(\alpha + x_i^T\beta), \\
T_i &\sim \rm{Weibull}(a, m_i),
\end{align}

where $\alpha$ is an intercept and $\beta$ is a vector of coefficients. $\alpha$ and values of $\beta$ for the CGDB variables were estimated from the CGDB data. The probability that the coefficient of a binary variables was non-zero was $0.5$ and drawn from a Bernoulli distribution. Non-zero coefficients were drawn from a uniform distribution with lower bound of $-0.5$ and upper bound of $0.5$ so that the minimum and maximum hazard ratios were $0.61$ and $1.65$, respectively.

Time to right censoring was also sampled from a Weibull distribution but did not depend on covariates,

\begin{equation} 
U_i \sim \rm{Weibull}(\tilde{a}, \tilde{m}).
\end{equation}

Lastly, a two-part model was used to simulate time to study entry so that a proportion $\tilde{\pi}$ of patients could enter the study at $t=0$ while a proportion $1-\tilde{\pi}$ would enter at $t>0$. The probability of an entry time greater than $0$ was drawn from a Bernoulli distribution and positive entry dates were modeled using a lognormal distribution.

\begin{align}
d_i &\sim \rm{Bernoulli}(\tilde{\pi}) \\
V_i^{*} &\sim \rm{Lognormal}(\mu, \sigma^2) \\
V_i &=\begin{cases}
 0  \text{ if } d_i = 0\\
  V_i^{*}  \text{ if } d_i = 1
  \end{cases}
\end{align}

\subsection{Calibration} \label{subsec:sim-calibration}
The simulation was calibrated to be consistent with the CGDB NSCLC data. The Nelson-Aalen estimator of the cumulative hazard of overall survival was generally monotonically decreasing, which suggested that a Weibull distribution could adequately capture the baseline hazard. This was further supported by the similarity in fit between a spline-based survival model with one internal knot at the median of uncensored survival time and the Weibull model. As such, an intercept only Weibull model was used to estimate $a$ and $\alpha$ in the simulation. Similarly, a Weibull distribution was used to model time to right censoring (by reversing the censoring indicator in the model for overall survival. Both the model for overall survival and right censoring adjusted for left truncation.

Coefficients for the CGDB variables were estimated using a Cox proportional hazards model. The full coefficient vector combined the coefficients from the CGDFB variables with the coefficients from the simulated binary variables. The covariates were centered, $m = \alpha + x^T\beta - E(x^T\beta)$, so that predicted survival was consistent with the intercept only Weibull model.

The latent time to study entry is inherently unobserved, but we believe a two-part model can characterize the large fraction of tests in the observed data that occur near the diagnosis date  (\autoref{fig:left-trunc-time}). The probability of a positive entry time was set equal to $0.2$ and the lognormal distribution for positive entry times was derived using a right skewed distribution with a median value of $1$ and a mean value of $1.6$, both measured in years.  We found that this parameterization generated median survival times in the simulation that were similar to median survival in Kaplan-Meier estimates that did not adjust for left truncation in the CGDB.

\subsection{Implementation}
Datasets were simulated where each row contained information on a patient's observed survival time ($y_i)$, a censoring indicator ($e_i$), the time of study entry ($v_i$). 75\%  of patients were assigned to a training set and the remaining 25\% were assigned to a test set. Within the test set, Cox models with lasso penalties were fit among the observed (i.e., non-truncated) patients. Cross-validation with 10 folds was used to tune the  model and the value of $\lambda$ that minimized the partial likelihood deviance was selected. Models were fit with and without accounting for left truncation.

Predictions and model evaluations were performed on the test set. Testing was performed on both the "observed" sample, (non-truncated patients) and the "complete" sample (both truncated and non-truncated sample). The complete sample provides a true test of the method because there is non bias due to truncation.

This process was repeated $20$ times. For each iteration, models were evaluated based on a dataset consisting of $n=5,000$ patients. Simulations were peformed in both big and small $p$ scenarios. In the small $p$ scenario, $10$ binary variables were simulated in addition to the $11$ CGDB variables so that $p=21$; in the big $p$ scenario, $1,000$ binary variables were simulated so that $p=1,011$.

\subsection{Results}
We used calibration curves \citep{harrell2015regression} to compare models with and without a left truncation adjustment. Each point in the plot consists of simulated patients in a given decile of predicted survival at a given time point. The x-axis is the averaged predicted survival probability and the y-axis is the pseudo observed survival probability computed using the Kaplan-Meier estimator. Perfectly calibrated predictions are those that lie along the 45 degree line so that predicted and observed survival probabilities are equal. The plots consist of patients in the complete sample and are therefore a good test of the effectiveness of the left truncation adjustment.

Calibration curves are displayed in \autoref{fig:sim_coxlasso_calibration_complete} for both the big and small $p$ scenarios. Predicted survival probabilities from the model that does not adjust for left truncation are too high because the model is fit on the observed sample consisting only of patients that survived long enough to enter the sample. In other words, failing to adjust for left truncation results in an overestimation of survival probabilities. In contrast, the left-truncated adjusted model corrects for this bias. The small $p$ model is almost perfectly calibrated with points along the 45 degree line.

\begin{figure}[!ht]
\caption{Calibration of survival predictions for lasso model in simulation}
\label{fig:sim_coxlasso_calibration_complete}
\centering
\includegraphics[max size={.8\textwidth}{.8\textheight}]{figs/sim_coxlasso_calibration_complete.pdf} 
\begin{minipage}{\linewidth}
\footnotesize
Notes: The Cox model with lasso penalty using the training data and was subsequently used to predict the survival function for each patient in the test set. The small and big $p$ models contained $21$ and $1,011$ predictors, respectively. Patients were divided into deciles at each time point based on their predicted survival probabilities. Each point in the plot represents patients within a decile. ``Predicted survival probability" is the average of the predicted survival functions from the Cox model across patients within each decile and the ``Observed survival probability" is the Kaplan-Meier estimate of the survival function within each decile. A perfect prediction lies on the black 45 degree line. 
\end{minipage}
\end{figure}

The results of the big $p$ scenario are similar, although calibration is more difficult than in the lower dimensional setting with points lying farther from the 45 degree line on both extremes. Predictions tend to be more accurate for patients with predicted survival close to $0.5$, but underestimated for patients with higher survival probabilities and overestimated for patients with lower survival probabilities. Probability rescaling has been suggested in such cases (e.g. characteristic sigmoid-shaped calibration curves produced by different learning algorithms), for example with the use of Platt scaling \citep{platt1999scaling} or isotonic regression \citep{niculescu2005isotonic}.

A common metric used to evaluate survival models is the C-index \citep{harrell1996multivariable}, which provides a measure of how well a model will discriminate prognosis between patients. We computed the C-index using predictions from the Cox model with lasso penalty on high dimensional data. The results are presented in \autoref{tbl:sim-model-discrimination} and suggest that care should be taken when using the C-index to evaluate survival models in the presence of left-truncated data. When using the observed sample to evaluate the model, the C-index is higher in model that does not adjust for left truncation, even though it is poorly calibrated. Furthermore, even when using the complete sample, the C-indices are very similar.

\begin{table}[!ht]
\begin{center}
\begin{threeparttable}
\caption{Comparison of model discrimination in the simulation} \label{tbl:sim-model-discrimination}
\begin{tabularx}{.7\textwidth}{@{\extracolsep{\fill}}llr}
\hline
\multicolumn{1}{l}{Left truncation adjustment} & \multicolumn{1}{l}{Test sample} & \multicolumn{1}{l}{C-index} \\
\hline
\ExpandableInput{tables/sim_coxlasso_p21_cindex.txt}
\hline
\end{tabularx}
\scriptsize Cox models with lasso penalty were fit using the high-dimensional simulated data.
\end{threeparttable}
\end{center}
\end{table}

These results are perhaps not surprising. For example, when computing the C-index in the observed sample the model is evaluated in a test set that suffers from the same bias as the training set. Moreover, the similarity in the values of the C-index in the complete sample may be caused by the assumption of no direct correlation between the predictors and latent time to study entry. In other words, the bias in the coefficients is based only on the indirect correlation between the predictors and left truncation induced by (i) the correlation between the predictors and survival and (ii) the correlation between survival and the probability a patient is left-truncated. As we will see in \autoref{sec:rwd}, the impact of left truncation adjustment on the estimated values of the coefficients can be considerably larger if there is a more direct correlation.

\section{Real-world data application: Non-small cell lung cancer patients in the Clinico-Genomic Database (CGDB)} \label{sec:rwd}
We fit models using $\nPatients$ stage IV, NSCLC patients. 75\%  of patients were assigned to a training set and the remaining 25\% were assigned to a test set. The training set was then used to train and tune both small $p$ and large $p$ models with 13 and $\nCovsBig$ predictor variables, respectively. Both un-penalized and lasso penalized Cox models were fit, with and without a left truncation adjustment. In the penalized models, the optimal value of the shrinkage parameter,  $\lambda$, was selected using the value that minimized the partial likelihood deviance in 10-fold cross-validation.  The models were evaluated by assessing predictions on the test set.

In this real-world application, both small $p$ and large $p$ models were designed to highlight left truncation in a real-world dataset and its impact on inference, as opposed to optimizing predictive performance and the prognostic value of predictors. For this reason, models were kept non-complex and relatively uninformative, e.g. modeling linear and additive effects, applying no transformations to predictors (for instance, modeling mutations as binary), and including a variable, year of diagnosis, that is highly correlated with left truncation time but not necessarily with survival (\autoref{fig:left-trunc-time}). Survival was predicted from diagnosis. For an evaluation of meaningful prognostic clinical and genomic factors in NSCLC, see for example \citet{lai2020nsclc}.

Specifically, the smaller model consisted of the following variables: year of diagnosis, age at diagnosis, race, practice type (academic or community center), and the mutational status (mutated/not mutated) of 3 known prognostic biomarkers for NSCLC (TP53, KRAS, and EGFR). For each biomarker, 3 alteration types were considered: copy number (CN), short variant (SV), and rearrangement (RE). $\nTrainSmall$ patients remained in the training set after dropping patients with missing values on at least one covariate.

The larger model included all non-biomarker variables from the smaller model and added all mutations tested by Foundation Medicine. As the data come from both historic and current clinico-genomic testing products, the set of tested genes has changed over time. To resolve the resultant missing data problem, we constrained the mutation data only to columns with less than 50\% missing indicators and imputed the remaining mutation data using k nearest neighbors \citep{troyanskaya_missing_2001}, with k = 20.

\autoref{fig:hr-plot} displays hazard ratios for each model. The plot for the large $p$ only contains coefficients ranked in the top 10 by absolute value of the hazard ratio in either the left truncation adjusted or non-adjusted model.  The arrows denote the impact of adjusting for left truncation on the hazard ratio. In some cases the adjustment can have a large impact on the hazard ratios, often moving a large hazard ratio all the way to $1$ (i.e., to a coefficient of $0$). This is most notable with diagnosis year, which has a hazard ratio above 2 in both the small and large $p$ models but is set to $1$ after the left truncation adjustment. This effect is likely due to the strong negative correlation between calendar time and the probability that a patient is left truncated.

\begin{figure}[h]
\caption{Hazard ratios from Cox lasso model}
\label{fig:hr-plot}
\centering
\includegraphics[max size={\textwidth}{\textheight}]{figs/hr.pdf} 
\begin{minipage}{\linewidth}
\footnotesize
Notes: The figure for the "Large" $p$ model only includes variables ranked in the top 10 by the absolute value of the hazard ratio in either the left truncation adjusted or non-adjusted model.
\end{minipage}
\end{figure}

\autoref{tbl:model-discrimination}  and \autoref{fig:surv-calibration} evaluate the models based on predictions on the test set. \autoref{tbl:model-discrimination} compares models using the C-index. The C-index is relatively low in all of these illustrative models, though is artificially higher in the unadjusted models that do not account for left truncation - sometimes considerably so - and decreases after adjustment as the predictor hazard ratios are pushed toward $1$ and there are fewer differences in predicted relative risks across patients. As expected, the lasso model has a higher C-index in the larger $p$ setting as the Cox model is considerably overfitted.

\begin{table}[!ht]
\begin{center}
\begin{threeparttable}
\caption{Comparison of model discrimination in the CGDB} \label{tbl:model-discrimination}
\begin{tabularx}{\textwidth}{@{\extracolsep{\fill}}llrr}
\hline
\multicolumn{2}{l}{} & \multicolumn{2}{c}{C-index}  \\
\cmidrule(r){3-4}  
\multicolumn{1}{l}{Model} & \multicolumn{1}{l}{Covariate size} & \multicolumn{1}{l}{No adjustment} & \multicolumn{1}{l}{Adjustment}  \\
\hline
\ExpandableInput{tables/performance.txt}
\hline
\end{tabularx}
\scriptsize Note: ``Adjusted'' models incorporate left truncation times.
\end{threeparttable}
\end{center}
\end{table}

\autoref{fig:surv-calibration} displays the same survival calibration curves presented in the simulation section. As in the simulation study, the unadjusted models overpredict survival probabilities. In contrast, the adjusted model tends to underpredict survival, which contrasts from the well calibrated adjusted models from the simulations. Here, the adjusted models tend to be adequately calibrated for patients with lower survival probabilities but less well calibrated for patients with higher survival probabilities. One possible reason for this result is that the adjusted models pushed many of the hazard ratios to $1$ and do not adequately separate low and high risk patients.

\begin{figure}[h]
\caption{Calibration of survival predictions in the CGDB from the Cox lasso model}
\label{fig:surv-calibration}
\centering
\includegraphics[max size={\textwidth}{\textheight}]{figs/calibration_coxlasso.pdf} 
\end{figure}

\section{Discussion} \label{sec:discussion}
In this paper, we implement an efficient approach for a left truncation-adjusted, penalized Cox proportional hazards model and critically assess its impact on bias and inference in left truncated data. Using simulation studies and examples from real-world EHR and genomics data, we showed that there is a need for approaches that can both adjust for left truncation and model high dimensional data. In particular, our simulation shows that predictions from models that fail to adjust for left truncation will overestimate true survival probabilities whereas models that properly adjust can yield well calibrated survival predictions, even with high-dimensional data. Moreover, in our real-world use case, there were significant differences in coefficient estimates between models that did and did not adjust for left truncation.

Our results have important implications for specification, interpretation, and evaluation of prognostic survival models. Much of the difficulty arises because training and test sets are both subject to the same selection bias. In other words, a careful understanding of the data generating process is crucial and naive approaches can lead analysts astray. For example, our simulations suggest that commonly used C-index cannot differentiate between models that mimic the true data generating process and those that do not and may therefore lead to misleading conclusions in the presence of left truncated data. On the other hand, metrics that focus on predicted survival probabilities such as calibration curves are better able to determine whether a model can accurately predict the risk of an event.

In our real-world use case, an analyst could have mistakenly concluded that the high C-index in the non-adjusted model was indicative of good model performance when it was in fact driven primarily by a high correlation between predictors and left truncation. For instance, in our large $p$ model the hazard ratio for the predictor "year of diagnosis" moved from being the largest in absolute value to $1$ after adjusting for left truncation. The high C-index in the non-adjusted model was artificial because patients diagnosed earlier appeared to survive longer due to immortal time bias.

Although not the primary focus of this paper, our approach generated findings that were both consistent with prior literature in genomics and clinically plausible. One of the mutation features that had the highest importance was the presence of short variants in the gene KEAP1. This finding is supported by similar reports of risk-conferring status in an independent lung cancer cohort \citep{shen_harnessing_2019}. Conversely, variants in EGFR and ALK were found to be protective in the CGDB data, putatively because mutations in these genes qualify patients to receive targeted therapies and therefore benefit from improved response and longer survival.

We conclude with several recommendations. When making predictions with survival data is of interest, it is critical to define the risk set appropriately. To do this requires recognizing the underlying process by which individuals are captured in the data in the first place; i.e., when and how individuals came to be included in the database of interest. One of the motivations for the current work derives from the advent of biobank and genomic test databases that store information on patient illness - presumably after the patient has been diagnosed and ``at risk'' for some time. In these data, overlooking this ``delayed entry'' of at-risk patients would lead to adverse consequences including misleading conclusions, as we demonstrate in this work. In addition, we recommend that models should be evaluated using both discrimination and calibration based metrics.

The implementation of this approach is available in \pkg{glmnet} v.X.X.X on CRAN.

\pdfbookmark[1]{References}{References}
\bibliography{references}

\end{document}

### 2021-01-04 R/run_sim.R
Increase number of Monte Carlo replications from 40 to 200

@@ -688,7 +688,7 @@ run_sim <- function(n_sims = 1, x, params,
                              "calibrate.list", "calibrate.survfit", "run_sim1"),
                  .packages = c("data.table", "dplyr"), # Used in functions without namespaces (pipes for dplyr)
                  .combine = "combine_parallel",
-                 .errorhandling = "pass",
+                 .errorhandling = "remove",
                  .init = rep(list(list()), 5)
   ) %dorng% {
     if (i %% 5 == 0) print(i)

### 2021-01-04 main.R

@@ -36,7 +36,7 @@ source("R/calibrate.R")
 set.seed(77)
 theme_set(theme_bw())
 center_title <- function() theme(plot.title = element_text(hjust = 0.5))
-N_SIMS <- 50
+N_SIMS <- 200
 
 # Caching
 RERUN_CACHE <- TRUE # Set to TRUE to rerun all cached results
@@ -47,7 +47,7 @@ if (!dir.exists("cache")){
 # Parallel
 PARALLEL <- TRUE
 if (PARALLEL) {
-  cl <- parallel::makeCluster(24, setup_strategy = "sequential", outfile = "simout")
+  cl <- parallel::makeCluster(20, setup_strategy = "sequential", outfile = "simout")
   registerDoParallel(cl)
 }
 

### 2021-01-05 R/run_sim.R
Better error handling for simulation

@@ -631,8 +631,11 @@ plot_calibration_sim <- function() {
     theme(legend.position = "bottom") 
 }
 
-run_sim1 <- function(data, method = "coxph", parallel = FALSE,
+run_sim1 <- function(x, params, method = "coxph", parallel = FALSE,
                      alpha = 1, lambda = NULL) {
+  # Data 
+  data <- sim_survdata(X = x, params = params)
+  
   # Modeling
   train_test <- make_train_test(data)
   fits <- fit_models(train = train_test$train, method = method,
@@ -688,13 +691,25 @@ run_sim <- function(n_sims = 1, x, params,
                              "calibrate.list", "calibrate.survfit", "run_sim1"),
                  .packages = c("data.table", "dplyr"), # Used in functions without namespaces (pipes for dplyr)
                  .combine = "combine_parallel",
-                 .errorhandling = "remove",
+                 .errorhandling = "pass",
                  .init = rep(list(list()), 5)
   ) %dorng% {
     if (i %% 5 == 0) print(i)
-    data <- sim_survdata(X = x, params = params)
-    run_sim1(data = data, method = method, parallel = parallel,
-             alpha = alpha, lambda = lambda)
+    it <- 1
+    while (it > 0) {
+      tryout <- try(run_sim1(x = x, params = params, method = method, parallel = parallel,
+                             alpha = alpha, lambda = lambda))
+      if (inherits(tryout, "try-error")) {
+        print(paste0("Error during iteration ", i,". Rerunning..."))
+        if (it == 5) stop(paste0("Too many errors during iteration ", i), 
+                          call. = FALSE)
+        it <- it + 1
+        tryout <- NULL
+      } else{
+        break
+      }
+    }
+    return(tryout)
   }
   run_time <- proc.time() - ptm
 
@@ -708,9 +723,9 @@ run_sim <- function(n_sims = 1, x, params,
     calibration = rbindlist(out$calibration, idcol = "sim"),
     run_time = run_time,
     n_patients = nrow(x),
-    p = ncol(x),
-    n_sims = n_sims
+    p = ncol(x)
   )
+  res$n_sims <- length(unique(res$surv_data$sim))
   return(res)
 }
 

### 2021-01-05 main.R
add package installation step

@@ -1,6 +1,11 @@
 # Setup ------------------------------------------------------------------------
 rm(list = ls())
 
+# Installation
+install.packages("BiocManager") # to install Bioconductor manager
+BiocManager::install("impute") # to install impute from Bioconductor
+renv::restore() # to install all other dependencies
+
 # Packages
 library("data.table")
 library("doParallel")

### 2021-01-05 main.R
move installation to README

@@ -1,11 +1,6 @@
 # Setup ------------------------------------------------------------------------
 rm(list = ls())
 
-# Installation
-install.packages("BiocManager") # to install Bioconductor manager
-BiocManager::install("impute") # to install impute from Bioconductor
-renv::restore() # to install all other dependencies
-
 # Packages
 library("data.table")
 library("doParallel")

### 2021-01-05 manuscript.tex
new edits

@@
 \begin{abstract}
-    High-dimensional data are common in the medical field as high-throughput screening, detailed electronic health record (EHR) information, and comprehensive genomics testing are now used to capture the patient profile.
+    High-dimensional data are becoming increasingly common in the medical field as large volumes of patient information are collected and processed by high-thoroughput screening, electronic health records (EHRs), and comprehensive genomic testing.
 Statistical models that attempt to study the effects of many predictors on survival typically implement feature selection or penalized methods to mitigate the undesirable consequences of overfitting.
@@
 \section{Introduction}
-The recent development and accessibility of genomic (DNA sequencing, microarrays), proteomic (immuno-histology, tissue arrays), digital (wearables), and imaging (PET, fMRI) technologies have led to an increasing complexity and dimensionality of patient health information.
-The level of person-level and disease-specific detail captured by these technologies offers an exciting new opportunity to understand patient history and prognosis by combining these molecular data with macro-scale clinical data on patient characteristics, treatment, and outcomes.
+The recent development and accessibility of genomic (DNA sequencing, microarrays), proteomic (immuno-histology, tissue arrays), clinical (electronic health records), digital (wearables), and imaging (PET, fMRI) technologies have led to an increasing complexity and dimensionality of patient health information.
+The amount of person-level and disease-specific detail captured by these technologies offers an exciting new opportunity to understand patient history and prognosis by combining these molecular data with macro-scale clinical data on patient characteristics, treatment, and outcomes.
 Specifically, these data may offer insights into risk factors affecting patient-related outcomes leading to interest in developing prognostic tools to improve clinical decision-making and patient care \citep{gui2005penalized, wishart2010predict, ow2016big, yousefi2017predicting}.
@@
 In a high-dimensional setting, computationally efficient regularization and shrinkage techniques have been developed \citep{tibshirani1997lasso, friedman2010regularization, simon2011regularization} for the Cox model.
-Similar to ordinary least squares regression (OLS), these penalized regression methods estimate regression coefficients by minimizing the sum of squared errors, but place a constraint in the equation to minimize the sum of coefficients.
+Similar to ordinary least squares (OLS) regression, these penalized models estimate regression coefficients by minimizing the sum of squared errors, but place a constraint in the equation to minimize the sum of coefficients.
 This constraint (controlled by a parameter, $\lambda$) penalizes a high number of variables in the model and shrinks the coefficients of less influential features towards zero - acting, in cases when some coefficients are shrunk fully to zero, as automatic feature selection.
@@
 An analysis of the observed patients is subject to an immortal time bias because observed patients cannot die prior to entering the study cohort \citep{levesque2010problem, giobbie2013challenges}.
-An example in medicine is a dataset describing patients who receive a specific diagnostic or genetic test; patients who die before having the chance to be tested are excluded from the data so their survival times do not contribute to overall survival estimates.
+An example in medicine is a dataset describing patients who receive a specific diagnostic or genetic test; patients who die before having the chance to be tested are excluded from the data, and consequently their survival times do not contribute to overall survival estimates.
 Traditional survival models are inappropriate in this case and predicted survival probabilities are overestimated because of the exclusion of short survival times.
+Such bias has been observed in large oncology cohorts marked by a milestone entry, including patients who receive genomic tests \citep{kehl2020selectionbias} and specialized treatments \citep{newman2020immortaltime}.
+These databases are increasingly used to understand patient survival to support clinical decision-making, motivating the need to identify and address this survivor selection bias in practice.
 
@@
 \subsection{Left truncation}
-In the absence of left truncation, the risk set $R_i = \{ j : y_j \geq t_i \}$ assumes that every individual is at risk of an event at time 0 and continues to be at risk until the event occurs or the individual is censored.
-In this sense, all individuals enter the risk set at the same time and leave only at death or censoring.
+In the absence of left truncation, the risk set $R_i = \{ j : y_j \geq t_i \}$ assumes that every individual is at risk of an event at his or her respective time 0 and continues to be at risk until the event occurs or the individual is censored.
+In this sense, all individuals enter the risk set at the same time 0 and leave only at death or censoring.
 However, when survival times are left truncated, individuals are truly at risk prior to entering the study (e.g.
@@
 \begin{figure}[h]
-\caption{Patient follow-up.
-Left-truncated, right-censored patient follow-up in a hypothetical study cohort.
+\caption{Left-truncated, right-censored patient follow-up in a hypothetical study cohort, ordered chronologically by event time.
 Patients become eligible to enter the study after diagnosis (closed circle) and are followed until death (open circle) or censoring (cross).
@@
 Patients 2 and 5 (in red) are excluded from the analysis because of death and a right censoring event, respectively, prior to study eligibility.
-A naive analysis on this data would therefore be biased because it only contains survival time for ``super-survivors`` (those who have survived a long time) and the recently-diagnosed.
+A naive analysis on this data would therefore be biased because it only contains survival time for ``super-survivors`` (those who have survived a long time, such as patients 1 and 3) and the recently-diagnosed.
 
@@
 The left-truncated risk set $R_{i\ast} = \{ j : y_j \geq t_i > v_j \}$.
-In other words, subjects are only considered at risk in the analysis after they have been observed to be at risk (e.g.
+In other words, subjects are only considered at risk in the analysis after they have been observed to be at risk (i.e.
 exceeded their left truncation time).
 Here, event times $t_1 < t_2 < ...
-< t_m$ are ordered relative to the earliest chronological event time observed in the patient cohort, rather than assuming that all patients are at risk starting at the same referent time $0$.
+< t_m$ are ordered relative to the earliest chronological event time observed in the patient cohort (as shown in \autoref{fig:followup}), rather than assuming that all patients are at risk starting at the same referent time $0$.
 
@@
 We take a subset of $\nPatients$ stage IV, non-small cell lung cancer (NSCLC) patients in the CGDB as a use case for this study.
-The outcome of interest is survival time predicted from the date of Stage 4 diagnosis.
+The outcome of interest is survival time predicted from the date of stage IV diagnosis.
 Survival times are left-truncated because patients who die before receiving a Foundation Medicine test are logically not included in the cohort of Foundation Medicine test recipients.
 
-As shown in \autoref{fig:left-trunc-time} the distribution of left truncation times (time from diagnosis to genomic testing) is right-tailed with approximately $\entryMoreOneYear$ of patients diagnosed more than one year prior to receiving the Foundation Medicine genomic test.
-Moreover, the correlation between diagnosis year and left truncation time is strongly negative ({$\rho$} = $\corrDxYear$), suggesting the presence of left truncation survivor bias as patients with records in the CGDB diagnosed longer ago needed to survive more years before being tested.
+As shown in \autoref{fig:left-trunc-time}, the distribution of left truncation times (time from diagnosis to genomic testing) is right-tailed with approximately $\entryMoreOneYear$ of patients diagnosed more than one year prior to receiving the Foundation Medicine genomic test.
+Moreover, the correlation between diagnosis year and left truncation time is strongly negative ({$\rho$} = $\corrDxYear$), suggesting the presence of left truncation survivor bias as patients with records in the CGDB diagnosed longer ago needed to survive more years before the availability of genomic testing.
 
@@
 
-This process was repeated $20$ times.
+This process was repeated $200$ times.
 For each iteration, models were evaluated based on a dataset consisting of $n=5,000$ patients.
@@
 
-The results of the big $p$ scenario are similar, although calibration is more difficult than in the lower dimensional setting with points lying farther from the 45 degree line on both extremes.
-Predictions tend to be more accurate for patients with predicted survival close to $0.5$, but underestimated for patients with higher survival probabilities and overestimated for patients with lower survival probabilities.
-Probability rescaling has been suggested in such cases (e.g.
-characteristic sigmoid-shaped calibration curves produced by different learning algorithms), for example with the use of Platt scaling \citep{platt1999scaling} or isotonic regression \citep{niculescu2005isotonic}.
+The results of the big $p$ scenario are similar, although calibration is more difficult than in the lower dimensional setting with points lying farther from the 45 degree line for patients with higher survival probabilities.
+Predictions tend to be more accurate for patients with predicted survival close to and below $0.5$, but underestimated for patients with higher survival probabilities.
+Probability rescaling may be effective in such cases, for example with the use of Platt scaling \citep{platt1999scaling} or isotonic regression \citep{niculescu2005isotonic}.
 
@@
 These results are perhaps not surprising.
-For example, when computing the C-index in the observed sample the model is evaluated in a test set that suffers from the same bias as the training set.
+For example, when computing the C-index in the observed sample, the model is evaluated in a test set that suffers from the same bias as the training set.
 Moreover, the similarity in the values of the C-index in the complete sample may be caused by the assumption of no direct correlation between the predictors and latent time to study entry.
@@
 As the data come from both historic and current clinico-genomic testing products, the set of tested genes has changed over time.
-To resolve the resultant missing data problem, we constrained the mutation data only to columns with less than 50\% missing indicators and imputed the remaining mutation data using k nearest neighbors \citep{troyanskaya_missing_2001}, with k = 20.
+To resolve the resultant missing data problem, we constrained the mutation data only to biomarkers with less than 50\% missingness and imputed the remaining mutation data using k nearest neighbors \citep{troyanskaya_missing_2001}, with k = 20.
 
@@
 In some cases the adjustment can have a large impact on the hazard ratios, often moving a large hazard ratio all the way to $1$ (i.e., to a coefficient of $0$).
-This is most notable with diagnosis year, which has a hazard ratio above 2 in both the small and large $p$ models but is set to $1$ after the left truncation adjustment.
+This is most notable with diagnosis year, which has a hazard ratio of approximately $1.5$ in both the small and large $p$ models but is set to $1$ after the left truncation adjustment.
 This effect is likely due to the strong negative correlation between calendar time and the probability that a patient is left truncated.
@@
 One possible reason for this result is that the adjusted models pushed many of the hazard ratios to $1$ and do not adequately separate low and high risk patients.
+Probability rescaling such as Platt scaling or isotonic regression may be effective on the adjusted probabilities here, which resemble the characteristic sigmoid-shaped calibration curves produced by different learning algorithms, including the lasso.
 
@@
 \section{Discussion} \label{sec:discussion}
-In this paper, we implement an efficient approach for a left truncation-adjusted, penalized Cox proportional hazards model and critically assess its impact on bias and inference in left truncated data.
+In this paper, we implement an efficient approach for a left truncation-adjusted, penalized Cox proportional hazards model and critically examine its impact on bias and inference in left truncated data.
 Using simulation studies and examples from real-world EHR and genomics data, we showed that there is a need for approaches that can both adjust for left truncation and model high dimensional data.
 In particular, our simulation shows that predictions from models that fail to adjust for left truncation will overestimate true survival probabilities whereas models that properly adjust can yield well calibrated survival predictions, even with high-dimensional data.
-Moreover, in our real-world use case, there were significant differences in coefficient estimates between models that did and did not adjust for left truncation.
+Moreover, in our real-world use case, there were significant differences in coefficient estimates between models that did and did not adjust for left truncation, compromising inference.
 
@@
 In other words, a careful understanding of the data generating process is crucial and naive approaches can lead analysts astray.
-For example, our simulations suggest that commonly used C-index cannot differentiate between models that mimic the true data generating process and those that do not and may therefore lead to misleading conclusions in the presence of left truncated data.
+For example, our simulations suggest that commonly used C-index cannot differentiate between models that mimic the true data generating process and those that do not, and may therefore lead to misleading conclusions in the presence of left truncated data.
 On the other hand, metrics that focus on predicted survival probabilities such as calibration curves are better able to determine whether a model can accurately predict the risk of an event.
@@
 In these data, overlooking this ``delayed entry'' of at-risk patients would lead to adverse consequences including misleading conclusions, as we demonstrate in this work.
-In addition, we recommend that models should be evaluated using both discrimination and calibration based metrics.
+Additionally, patient cohorts marked by a milestone entry, such as genomic testing, may not only be subject to immortal time bias but also to a temporal selection bias.
+The latter may occur if left truncation time is not quasi-independent of survival time, for example, if genomic testing is ordered selectively for patients with worse prognosis.
+This was uncovered in a cohort of multi-stage cancer patients who received genomic testing \citep{kehl2020selectionbias} and can affect inference even in the presence of a left truncation adjustment.
+While the cohort of lung cancer patients in our real-world use case was restricted to stage IV diagnoses, mitigating this potential selection bias \citep{kehl2020selectionbias}, there exist formal tests to assess quasi-independence via the conditional Kendall's tau \citep{martin2005tau}.
+Finally, we recommend that models should be evaluated using both discrimination and calibration based metrics.
 

### 2021-01-05 R/plot_patient_followup.R
manuscript says patient 5 is censored- change to reflect this

@@ -4,10 +4,10 @@
 #' @return A [ggplot2][ggplot2::ggplot2] object.
 #' @export
 plot_patient_followup <- function(){
-  dat <- data.frame(ID = 1:7, event = c(0, 1, 1, 1, 1, 0, 0), 
+  dat <- data.frame(ID = 1:7, event = c(0, 1, 1, 1, 0, 0, 0), 
                     t1 = c(1989, 1991, 1993, 2007, 2008, 2012, 2014), 
                     t2 = c(2019, 1998, 2013, 2017, 2009, 2019, 2019), 
-                    censored = c(1, 0, 0, 0, 0, 1, 1)) #, .Names = c("ID", "Death", "t1", "t2", "Censored")) #, class = "data.frame", row.names = c(NA, -6L))
+                    censored = c(1, 0, 0, 0, 1, 1, 1)) #, .Names = c("ID", "Death", "t1", "t2", "Censored")) #, class = "data.frame", row.names = c(NA, -6L))
   
   # Create event variable
   #dat$Death <- with(dat, ifelse(Death, "Dead", "Censored"))

### 2021-01-06 manuscript.tex
Minor edits to manuscript text

@@
 %\usepackage{pgfplots}
-%\usepackage[font={bf}]{caption}
 \usepackage{latexsym}
@@
 Statistical models that attempt to study the effects of many predictors on survival typically implement feature selection or penalized methods to mitigate the undesirable consequences of overfitting.
-In some cases survival data is also left truncated which can create an immortal time bias, but penalized survival methods that adjust for left truncation are unavailable in common software.
+In some cases survival data is also left-truncated which can create an immortal time bias, but penalized survival methods that adjust for left truncation are unavailable in common software.
 To address these challenges, we implement a penalized Cox proportional hazards model for left-truncated and right-censored survival data and assess implications of left truncation adjustment on bias and inference.
 We use simulation studies and examples from high-dimensional real-world EHR and genomics data to highlight the pitfalls of failing to account for left truncation in survival modeling.
-The extension of survival modeling to left truncated data is available in the latest release of \pkg{glmnet}.
+The extension of survival modeling to left-truncated data is available in the latest release of \pkg{glmnet}.
 \end{abstract}
@@
 diseased) for some time so that patients with shorter survival times are unobserved.
-In other words, the data are left truncated because patients are only included if they survive beyond some initial milestone that marks entry into the observed data.
+In other words, the data are left-truncated because patients are only included if they survive beyond some initial milestone that marks entry into the observed data.
 An analysis of the observed patients is subject to an immortal time bias because observed patients cannot die prior to entering the study cohort \citep{levesque2010problem, giobbie2013challenges}.
@@
 
-Here, we implement a penalized Cox proportional hazards for left-truncated and right-censored survival data and critically assess scenarios in which bias arises from and is exacerbated by left truncation.
+Here, we implement a penalized Cox proportional hazards model for left-truncated and right-censored (LTRC) survival data and critically assess scenarios in which bias arises from and is exacerbated by left truncation.
 We apply the method to simulated data and a real-world use case from the Flatiron Health and Foundation Medicine Clinico-Genomic Database (CGDB).
 Our results highlight the importance of adjusting for left truncation for both inference and prediction.
-Furthermore, we show that use of inappropriate metrics for model evaluation can result in overly optimistic conclusions about model performance.
-For implementation of the proposed method, we introduce the left truncation extension to survival modeling in the R package \pkg{glmnet} v.X.X (CITE LATEST PACKAGE).
+Furthermore, we show that use of inappropriate metrics for model evaluation can result in overly optimistic conclusions about the performance of predictive models.
+We implementation of the proposed method, we use the left truncation extension to survival modeling introduce in v.X.X of the \textsf{R} package \pkg{glmnet} (CITE LATEST PACKAGE).
 
-The remainder of this paper is organized as follows: In \autoref{sec:methods}, we introduce the penalized regression and left truncation framework and present the efficient method that adjusts for both.
-\autoref{sec:simulation} presents the simulation results and in \autoref{sec:rwd} the method is applied to a real-world dataset combining electronic health records and genomic testing.
+The remainder of this paper is organized as follows: In \autoref{sec:methods}, we introduce the penalized regression and left truncation framework and present the efficient method that incorporates both.
+\autoref{sec:simulation} presents the simulation results and \autoref{sec:rwd} considers a real-world dataset that combines electronic health records and genomic testing information.
 Finally, we discuss general takeaways and survival modeling recommendations in \autoref{sec:discussion}.
@@
 For each patient $i$, we have data of the form $y_i$, $e_i$, $x_i$ where $y_i$ is the observed survival time, defined as the time between baseline and event, $e_i$ is the event or failure of interest (e.g.
-death), and $x_i = (x_{i,1}, x_{i,2} \ldots, x_{i,p})$ is a vector of covariates of length $p$ used in the regression model.
+death), and $x_i = (x_{i,1}, x_{i,2} \ldots, x_{i,p})$ is a vector of predictors of length $p$ used in the regression model.
 The event or failure of interest $e_i$ is 1 if the event is observed or 0 if the event is right-censored.
@@
 In this sense, all individuals enter the risk set at the same time 0 and leave only at death or censoring.
-However, when survival times are left truncated, individuals are truly at risk prior to entering the study (e.g.
+However, when survival times are left-truncated, individuals are truly at risk prior to entering the study (e.g.
 prior to being observed for follow-up).
@@
 \begin{figure}[h]
-\caption{Left-truncated, right-censored patient follow-up in a hypothetical study cohort, ordered chronologically by event time.
-Patients become eligible to enter the study after diagnosis (closed circle) and are followed until death (open circle) or censoring (cross).
-However, patients who die before the start of the study (dashed blue line) are left truncated (in red), and only those who have survived until the study start (in black) are observed.}
-\label{fig:followup}
 \centering
 \includegraphics[max size={\textwidth}{\textheight}]{figs/followup.pdf} 
+\caption{Left-truncated and right-censored patient follow-up in a hypothetical study cohort, ordered chronologically by event time.
+Patients become eligible to enter the study after diagnosis (closed circle) and are followed until death (open circle) or censoring (cross).
+However, patients who die before the start of the study (dashed blue line) are left-truncated (in red), and only those who have survived until the study start (in black) are observed.}
+\label{fig:followup}
 \end{figure}
@@
 Survival is defined as the time from diagnosis until death, but the study does not begin until 2011 (dashed blue vertical line).
-Patients 2 and 5 (in red) are excluded from the analysis because of death and a right censoring event, respectively, prior to study eligibility.
-A naive analysis on this data would therefore be biased because it only contains survival time for ``super-survivors`` (those who have survived a long time, such as patients 1 and 3) and the recently-diagnosed.
+Patients 2 and 5 (in red) are excluded from the analysis because of death and right censoring, respectively, prior to study eligibility.
+A naive analysis on this data would therefore be biased because it only contains survival time for ``super-survivors'' (those who have survived a long time, such as patients 1 and 3) and the recently-diagnosed.
 
@@
 \subsection{Penalized regression with left-truncated and right-censored data}
-Penalized regression methods such as $l_1$ (lasso) \citep{tibshirani1996regression} and $l_2$ (ridge) \citep{tikhonov1963ridge} regression, elastic net regression \citep{zou2005regularization}, and SCAD (Smoothly Clipped Absolute Deviation) \citep{xie2009scad} have been developed to overcome the limitation of fitting a model to a large number of covariates.
+Penalized regression methods such as $l_1$ (lasso) \citep{tibshirani1996regression} and $l_2$ (ridge) \citep{tikhonov1963ridge} regression, elastic net regression \citep{zou2005regularization}, and SCAD (Smoothly Clipped Absolute Deviation) \citep{xie2009scad} have been developed to overcome the limitation of fitting a model to a large number of predictors.
 In a high dimensional feature space, where the number of predictors is larger than the sample size (i.e., $p$ > $n$), a model may exhibit low bias but suffer from poor prediction accuracy and limited generalizability, a consequence of overfitting the data.
@@
 Applying the standard penalized Cox implementation in the software \pkg{glmnet} to left-truncated survival data will result in biased estimates.
-However, the extension to adjust for left truncation is straightforward.
-Given survival data with left truncation and right censoring, we solve \autoref{eqn:simon} with a modification to the risk set $R_i$:
+However, an extension to LTRC data is straightforward.
+We solve for \autoref{eqn:simon} with a modification to the risk set $R_i$:
 
@@
 
-where the left-truncated risk set $R_{i\ast} = \{ j : y_j \geq t_i > v_j \}$; that is, it considers subjects at risk only after their left truncation times.
-The method can take lasso, ridge, and elastic net penalties for $\lambda P_{\alpha}\beta$ defined elsewhere \citep{friedman2010regularization} and is a simple extension of default methods in \pkg{glmnet}.
-The efficient implementation of this method is non-trivial, as the usual updating of risk set components cannot be used when left truncation is present.
+where the left-truncated risk set $R_{i\ast} = \{ j : y_j \geq t_i > v_j \}$; that is, subjects are only considered at risk after their left truncation times.
+An efficient implementation of this method is non-trivial, as the usual updating of risk set components cannot be used when left truncation is present.
 This implementation is available in the release of \pkg{glmnet} v.
@@
 Patients in the CGDB have received at least one Foundation Medicine tumor genomic test as well as have had their electronic health records captured by Flatiron Health.
-Together, these data sources represent thousands of potential predictors per patient, ranging from demographic, lab, and treatment information to the specific mutations detected on each biomarker included in a Foundation Medicine baitset.
+Together, these data sources represent thousands of potential predictors per patient, ranging from demographic, laboratory, and treatment information to the specific mutations detected on each biomarker included in a Foundation Medicine baitset.
 
@@
 \begin{figure}[h]
+ \centering
+\includegraphics[max size={\textwidth}]{figs/left_trunc_hist.pdf} 
 \caption{Distribution of left truncation time (days) in non-small cell lung cancer patients in the CGDB}
  \label{fig:left-trunc-time}
- \centering
-\includegraphics[max size={\textwidth}{\textheight}]{figs/left_trunc_hist.pdf} 
 \end{figure}
@@
 \section{Simulation study} \label{sec:simulation}
-Simulations were conducted to assess the performance of the method with left truncated survival data.
+Simulations were conducted to assess the performance of the method with LTRC data.
 To set notation, let $T_i$, $U_i$, and $V_i$ be latent survival, right censoring, and study entry times for patient $i$, respectively.
 The simulation proceeds by simulating $T_i$, $U_i$, and $V_i$ from suitable survival distributions and assumes that $T_i$, $U_i$, and $V_i$ are uncorrelated.
-The observed survival time is given by $Y_i = \rm{min}(T_i, U_i)$ and a patient is right-censored if $T_i > U_i$.
+The observed survival time is given by $Y_i = \min(T_i, U_i)$ and a patient is right-censored if $T_i > U_i$.
 Patients are left-truncated (i.e., hidden from the sample) if $V_i > Y_i$.
@@
 
-Latent survival time was simulated from a proportional hazards Weibull distribution with shape parameter $a$ and scaled parameter $m_i$ based on evidence from the CGDB as detailed in \autoref{subsec:sim-calibration}.
-Covariates predict latent survival time through their effect on the scale parameter,
+Latent survival time was simulated from a proportional hazards Weibull distribution with shape parameter $a$ and scale parameter $m_i$ based on evidence from the CGDB as detailed in \autoref{subsec:sim-calibration}.
+Predictors predict latent survival time through their effect on the scale parameter,
 
@@
 The probability that the coefficient of a binary variables was non-zero was $0.5$ and drawn from a Bernoulli distribution.
-Non-zero coefficients were drawn from a uniform distribution with lower bound of $-0.5$ and upper bound of $0.5$ so that the minimum and maximum hazard ratios were $0.61$ and $1.65$, respectively.
+Non-zero coefficients were drawn from a uniform distribution with lower bound of $-0.25$ and upper bound of $0.25$ so that the minimum and maximum hazard ratios were $0.78$ and $1.28$, respectively.
 
-Time to right censoring was also sampled from a Weibull distribution but did not depend on covariates,
+Time to right censoring was also sampled from a Weibull distribution but did not depend on predictors,
 
@@
 
-Coefficients for the CGDB variables were estimated using a Cox proportional hazards model.
-The full coefficient vector combined the coefficients from the CGDFB variables with the coefficients from the simulated binary variables.
-The covariates were centered, $m = \alpha + x^T\beta - E(x^T\beta)$, so that predicted survival was consistent with the intercept only Weibull model.
+The CGDB variables were standardized to have mean $0$ and standard deviation $1$ and coefficients were estimated using a Cox proportional hazards model.
+The full coefficient vector combined the coefficients from the CGDB variables with the coefficients from the simulated binary variables.
+The scale parameter was set to $m = \alpha + x^T\beta - E(x^T\beta)$ so that predicted survival averaged across patients was equal to predicted survival from the intercept only Weibull model.
 
-The latent time to study entry is inherently unobserved, but we believe a two-part model can characterize the large fraction of tests in the observed data that occur near the diagnosis date  (\autoref{fig:left-trunc-time}).
+Latent time to study entry is inherently unobserved, but we believe a two-part model can characterize the large fraction of tests in the observed data that occur near the diagnosis date  (\autoref{fig:left-trunc-time}).
 The probability of a positive entry time was set equal to $0.2$ and the lognormal distribution for positive entry times was derived using a right skewed distribution with a median value of $1$ and a mean value of $1.6$, both measured in years.
@@
 75\%  of patients were assigned to a training set and the remaining 25\% were assigned to a test set.
-Within the test set, Cox models with lasso penalties were fit among the observed (i.e., non-truncated) patients.
+
+Within the training set, Cox models with lasso penalties were fit among the ``observed'' (i.e., non-truncated) patients.
 Cross-validation with 10 folds was used to tune the  model and the value of $\lambda$ that minimized the partial likelihood deviance was selected.
-Models were fit with and without accounting for left truncation.
+Models were fit with and without adjusting for left truncation.
+All models adjusted for right censoring.
 
 Predictions and model evaluations were performed on the test set.
-Testing was performed on both the "observed" sample, (non-truncated patients) and the "complete" sample (both truncated and non-truncated sample).
-The complete sample provides a true test of the method because there is non bias due to truncation.
+The models were evaluated using the ``complete'' sample consisting of both truncated and non-truncated patients.
+The use of the complete sample for model evaluation is a key advantage of a simulation study since it is unobservable in real-world applications.
 
@@
 For each iteration, models were evaluated based on a dataset consisting of $n=5,000$ patients.
-Simulations were peformed in both big and small $p$ scenarios.
-In the small $p$ scenario, $10$ binary variables were simulated in addition to the $11$ CGDB variables so that $p=21$; in the big $p$ scenario, $1,000$ binary variables were simulated so that $p=1,011$.
+Simulations were performed in both big and small $p$ scenarios.
+In the small $p$ scenario, $10$ binary variables were simulated in addition to the $11$ CGDB variables so that $p=21$; in the large $p$ scenario, $1,000$ binary variables were simulated so that $p=1,011$.
 
@@
 
-Calibration curves are displayed in \autoref{fig:sim_coxlasso_calibration_complete} for both the big and small $p$ scenarios.
+Calibration curves are displayed in \autoref{fig:sim_coxlasso_calibration_complete} for both the small and large $p$ scenarios.
 Predicted survival probabilities from the model that does not adjust for left truncation are too high because the model is fit on the observed sample consisting only of patients that survived long enough to enter the sample.
@@
 \begin{figure}[!ht]
+\centering
+\includegraphics[max size={.8\textwidth}{.8\textheight}]{figs/sim_coxlasso_calibration_complete.pdf} 
 \caption{Calibration of survival predictions for lasso model in simulation}
 \label{fig:sim_coxlasso_calibration_complete}
-\centering
-\includegraphics[max size={.8\textwidth}{.8\textheight}]{figs/sim_coxlasso_calibration_complete.pdf} 
 \begin{minipage}{\linewidth}
@@
 Notes: The Cox model with lasso penalty using the training data and was subsequently used to predict the survival function for each patient in the test set.
-The small and big $p$ models contained $21$ and $1,011$ predictors, respectively.
+The small and large $p$ models contained $21$ and $1,011$ predictors, respectively.
 Patients were divided into deciles at each time point based on their predicted survival probabilities.
 Each point in the plot represents patients within a decile.
-``Predicted survival probability" is the average of the predicted survival functions from the Cox model across patients within each decile and the ``Observed survival probability" is the Kaplan-Meier estimate of the survival function within each decile.
+The ``Predicted survival probability" is the average of the predicted survival probabilities from the Cox model across patients within each decile and the ``Observed survival probability" is the Kaplan-Meier estimate of the proportion surviving within each decile.
 A perfect prediction lies on the black 45 degree line.
@@
 
-The results of the big $p$ scenario are similar, although calibration is more difficult than in the lower dimensional setting with points lying farther from the 45 degree line for patients with higher survival probabilities.
-Predictions tend to be more accurate for patients with predicted survival close to and below $0.5$, but underestimated for patients with higher survival probabilities.
+The results of the large $p$ scenario are similar, although calibration is more difficult than in the lower dimensional setting with points lying farther from the 45 degree line for patients with higher survival probabilities.
+Predictions tend to be more accurate for patients with predicted survival close to and below $0.5$, but slightly underestimated for patients with higher survival probabilities.
 Probability rescaling may be effective in such cases, for example with the use of Platt scaling \citep{platt1999scaling} or isotonic regression \citep{niculescu2005isotonic}.
@@
 The results are presented in \autoref{tbl:sim-model-discrimination} and suggest that care should be taken when using the C-index to evaluate survival models in the presence of left-truncated data.
-When using the observed sample to evaluate the model, the C-index is higher in model that does not adjust for left truncation, even though it is poorly calibrated.
+When using the observed sample to evaluate the model, the C-index is higher in a model that does not adjust for left truncation, even though it is poorly calibrated.
 Furthermore, even when using the complete sample, the C-indices are very similar.
@@
 In other words, the bias in the coefficients is based only on the indirect correlation between the predictors and left truncation induced by (i) the correlation between the predictors and survival and (ii) the correlation between survival and the probability a patient is left-truncated.
-As we will see in \autoref{sec:rwd}, the impact of left truncation adjustment on the estimated values of the coefficients can be considerably larger if there is a more direct correlation.
+As we will see in \autoref{sec:rwd}, the impact of left truncation adjustment on the estimated values of the coefficients (and by extenstion the C-index) can be considerably larger if there is a more direct correlation.
 
@@
 75\%  of patients were assigned to a training set and the remaining 25\% were assigned to a test set.
-The training set was then used to train and tune both small $p$ and large $p$ models with 13 and $\nCovsBig$ predictor variables, respectively.
+The training set was then used to train and tune both small $p$ and large $p$ models with $11$ and $\nCovsLarge$ predictor variables, respectively.
 Both un-penalized and lasso penalized Cox models were fit, with and without a left truncation adjustment.
@@
 
-In this real-world application, both small $p$ and large $p$ models were designed to highlight left truncation in a real-world dataset and its impact on inference, as opposed to optimizing predictive performance and the prognostic value of predictors.
-For this reason, models were kept non-complex and relatively uninformative, e.g.
+Both the small and large $p$ models were designed to highlight left truncation in a real-world dataset and its impact on inference, as opposed to optimizing predictive performance and the prognostic value of predictors.
+For this reason, model specifications were relatively simple, e.g.
 modeling linear and additive effects, applying no transformations to predictors (for instance, modeling mutations as binary), and including a variable, year of diagnosis, that is highly correlated with left truncation time but not necessarily with survival (\autoref{fig:left-trunc-time}).
@@
 
-Specifically, the smaller model consisted of the following variables: year of diagnosis, age at diagnosis, race, practice type (academic or community center), and the mutational status (mutated/not mutated) of 3 known prognostic biomarkers for NSCLC (TP53, KRAS, and EGFR).
-For each biomarker, 3 alteration types were considered: copy number (CN), short variant (SV), and rearrangement (RE).
+The smaller model consisted of the following variables: year of diagnosis, age at diagnosis, race, practice type (academic or community center), and the mutational status (mutated/not mutated) of 3 known prognostic biomarkers for NSCLC (TP53, KRAS, and EGFR).
+Short variants (SV) were considered for each biomarker while copy number alterations (CN) and rearrangements (RE) were also considered for TP53.
 $\nTrainSmall$ patients remained in the training set after dropping patients with missing values on at least one covariate.
 
-The larger model included all non-biomarker variables from the smaller model and added all mutations tested by Foundation Medicine.
+The larger model included all non-biomarker variables from the smaller model and added nearly all mutations tested by Foundation Medicine.
 As the data come from both historic and current clinico-genomic testing products, the set of tested genes has changed over time.
-To resolve the resultant missing data problem, we constrained the mutation data only to biomarkers with less than 50\% missingness and imputed the remaining mutation data using k nearest neighbors \citep{troyanskaya_missing_2001}, with k = 20.
+To resolve the resultant missing data problem, we constrained the mutation data to biomarkers with less than 30\% missingness and imputed the remaining mutation data using k nearest neighbors \citep{troyanskaya_missing_2001}, with k = 20.
+Biomarkers with zero variance or that were present in less than 0.1\% of patients were also excluded.
 
 \autoref{fig:hr-plot} displays hazard ratios for each model.
-The plot for the large $p$ only contains coefficients ranked in the top 10 by absolute value of the hazard ratio in either the left truncation adjusted or non-adjusted model.
+The plot for the large $p$ model only contains coefficients ranked in the top 10 by absolute value of the hazard ratio in either the left truncation adjusted or non-adjusted model.
 The arrows denote the impact of adjusting for left truncation on the hazard ratio.
-In some cases the adjustment can have a large impact on the hazard ratios, often moving a large hazard ratio all the way to $1$ (i.e., to a coefficient of $0$).
+In some cases the adjustment can have a considerable impact on the hazard ratios, often moving a large hazard ratio all the way to $1$ (i.e., to a coefficient of $0$).
 This is most notable with diagnosis year, which has a hazard ratio of approximately $1.5$ in both the small and large $p$ models but is set to $1$ after the left truncation adjustment.
-This effect is likely due to the strong negative correlation between calendar time and the probability that a patient is left truncated.
+This effect is likely due to the strong negative correlation between calendar time and the probability that a patient is left-truncated.
 
 \begin{figure}[h]
+\centering
+\includegraphics[max size={\textwidth}{\textheight}]{figs/hr.pdf} 
 \caption{Hazard ratios from Cox lasso model}
 \label{fig:hr-plot}
-\centering
-\includegraphics[max size={\textwidth}{\textheight}]{figs/hr.pdf} 
 \begin{minipage}{\linewidth}
@@
 \autoref{tbl:model-discrimination} compares models using the C-index.
-The C-index is relatively low in all of these illustrative models, though is artificially higher in the unadjusted models that do not account for left truncation - sometimes considerably so - and decreases after adjustment as the predictor hazard ratios are pushed toward $1$ and there are fewer differences in predicted relative risks across patients.
+The C-index is relatively low in all of these illustrative models, though is artificially higher in the unadjusted models that do not account for left truncation---sometimes considerably so---and decreases after adjustment as the predictor hazard ratios are pushed toward $1$ and there are fewer differences in predicted relative risks across patients.
 As expected, the lasso model has a higher C-index in the larger $p$ setting as the Cox model is considerably overfitted.
@@
 \autoref{fig:surv-calibration} displays the same survival calibration curves presented in the simulation section.
-As in the simulation study, the unadjusted models overpredict survival probabilities.
-In contrast, the adjusted model tends to underpredict survival, which contrasts from the well calibrated adjusted models from the simulations.
-Here, the adjusted models tend to be adequately calibrated for patients with lower survival probabilities but less well calibrated for patients with higher survival probabilities.
+As in the simulation study, the unadjusted models overpredict survival probabilities; however, in contrast to the simulations, the adjusted models are poorly calibrated.
+
+The adjusted models tend to be better calibrated for patients with the lowest survival probabilities but poorly calibrated for patients with higher survival probabilities.
 One possible reason for this result is that the adjusted models pushed many of the hazard ratios to $1$ and do not adequately separate low and high risk patients.
@@
 \begin{figure}[h]
+\centering
+\includegraphics[max size={\textwidth}{\textheight}]{figs/calibration_coxlasso.pdf} 
 \caption{Calibration of survival predictions in the CGDB from the Cox lasso model}
 \label{fig:surv-calibration}
-\centering
-\includegraphics[max size={\textwidth}{\textheight}]{figs/calibration_coxlasso.pdf} 
 \end{figure}
@@
 \section{Discussion} \label{sec:discussion}
-In this paper, we implement an efficient approach for a left truncation-adjusted, penalized Cox proportional hazards model and critically examine its impact on bias and inference in left truncated data.
+In this paper, we implemented an efficient approach for estimating penalized Cox proportional hazards model with LTRC survival data and compared it to penalized models designed for survival data that is right-censored but not left-truncated.
 Using simulation studies and examples from real-world EHR and genomics data, we showed that there is a need for approaches that can both adjust for left truncation and model high dimensional data.
@@
 In other words, a careful understanding of the data generating process is crucial and naive approaches can lead analysts astray.
-For example, our simulations suggest that commonly used C-index cannot differentiate between models that mimic the true data generating process and those that do not, and may therefore lead to misleading conclusions in the presence of left truncated data.
+For example, our simulations suggest that the commonly used C-index cannot differentiate between models that mimic the true data generating process and those that do not, and may therefore lead to misleading conclusions in the presence of LTRC data.
 On the other hand, metrics that focus on predicted survival probabilities such as calibration curves are better able to determine whether a model can accurately predict the risk of an event.
@@
 When making predictions with survival data is of interest, it is critical to define the risk set appropriately.
-To do this requires recognizing the underlying process by which individuals are captured in the data in the first place; i.e., when and how individuals came to be included in the database of interest.
+To do this requires recognizing the underlying process by which individuals are captured in the data in the first place; i.e., when and how individuals came to be included in the dataset of interest.
 One of the motivations for the current work derives from the advent of biobank and genomic test databases that store information on patient illness - presumably after the patient has been diagnosed and ``at risk'' for some time.
-In these data, overlooking this ``delayed entry'' of at-risk patients would lead to adverse consequences including misleading conclusions, as we demonstrate in this work.
+In these data, overlooking the ``delayed entry'' of at-risk patients could result in misleading conclusions, as we demonstrate in this work.
 Additionally, patient cohorts marked by a milestone entry, such as genomic testing, may not only be subject to immortal time bias but also to a temporal selection bias.
-The latter may occur if left truncation time is not quasi-independent of survival time, for example, if genomic testing is ordered selectively for patients with worse prognosis.
+The latter may occur if left truncation time is correlated with survival time, for example, if genomic testing is ordered selectively for patients with worse prognosis.
 This was uncovered in a cohort of multi-stage cancer patients who received genomic testing \citep{kehl2020selectionbias} and can affect inference even in the presence of a left truncation adjustment.
-While the cohort of lung cancer patients in our real-world use case was restricted to stage IV diagnoses, mitigating this potential selection bias \citep{kehl2020selectionbias}, there exist formal tests to assess quasi-independence via the conditional Kendall's tau \citep{martin2005tau}.
+While the cohort of lung cancer patients in our real-world use case was restricted to stage IV diagnoses, mitigating this potential selection bias \citep{kehl2020selectionbias}, it can be adjusted for by modeling the association between left truncation time and survival \citep{thiebaut2004choice, chiou2019transformation}.
 Finally, we recommend that models should be evaluated using both discrimination and calibration based metrics.

### 2021-01-06 manuscript.tex
An additional very minor edit
Typo fix

@@
 Furthermore, we show that use of inappropriate metrics for model evaluation can result in overly optimistic conclusions about the performance of predictive models.
-We implementation of the proposed method, we use the left truncation extension to survival modeling introduce in v.X.X of the \textsf{R} package \pkg{glmnet} (CITE LATEST PACKAGE).
+For implementation of the proposed method, we use the left truncation extension to survival modeling introduce in v.X.X of the \textsf{R} package \pkg{glmnet} (CITE LATEST PACKAGE).
 
@@
 For each iteration, models were evaluated based on a dataset consisting of $n=5,000$ patients.
-Simulations were performed in both big and small $p$ scenarios.
+Simulations were performed in both small and large $p$ scenarios.
 In the small $p$ scenario, $10$ binary variables were simulated in addition to the $11$ CGDB variables so that $p=21$; in the large $p$ scenario, $1,000$ binary variables were simulated so that $p=1,011$.

### 2021-01-06 manuscript.tex
Another typo fix

@@
 Furthermore, we show that use of inappropriate metrics for model evaluation can result in overly optimistic conclusions about the performance of predictive models.
-For implementation of the proposed method, we use the left truncation extension to survival modeling introduce in v.X.X of the \textsf{R} package \pkg{glmnet} (CITE LATEST PACKAGE).
+For implementation of the proposed method, we use the left truncation extension to survival modeling introduced in v.X.X of the \textsf{R} package \pkg{glmnet} (CITE LATEST PACKAGE).
 

### 2021-01-07 R/plot_patient_followup.R
reorder patient IDs

@@ -4,21 +4,24 @@
 #' @return A [ggplot2][ggplot2::ggplot2] object.
 #' @export
 plot_patient_followup <- function(){
-  dat <- data.frame(ID = 1:7, event = c(0, 1, 1, 1, 0, 0, 0), 
-                    t1 = c(1989, 1991, 1993, 2007, 2008, 2012, 2014), 
-                    t2 = c(2019, 1998, 2013, 2017, 2009, 2019, 2019), 
-                    censored = c(1, 0, 0, 0, 1, 1, 1)) #, .Names = c("ID", "Death", "t1", "t2", "Censored")) #, class = "data.frame", row.names = c(NA, -6L))
+  dat <- data.frame(ID = c(1:7), event = c(0, 1, 1, 1, 0, 0, 0),  # c(7, 1, 3, 4, 2, 6, 5)
+                    t1 = c(1991, 2008, 1993, 2007, 2012, 2014, 1989),
+                    t2 = c(1998, 2009, 2013, 2017, 2019, 2019, 2019), 
+                    #t1 = c(1989, 1991, 1993, 2007, 2008, 2012, 2014), 
+                   # t2 = c(2019, 1998, 2013, 2017, 2009, 2019, 2019), 
+                   censored = c(0, 1, 0, 0, 1, 1, 1))
+                   # censored = c(1, 0, 0, 0, 1, 1, 1)) #, .Names = c("ID", "Death", "t1", "t2", "Censored")) #, class = "data.frame", row.names = c(NA, -6L))
   
   # Create event variable
   #dat$Death <- with(dat, ifelse(Death, "Dead", "Censored"))
   
   # Create id.ordered, which is a factor that is ordered by t2
   # This will allow the plot to be ordered by increasing t2, if desired
-  dat$id.ordered <- factor(x = dat$ID, levels = order(dat$t2, decreasing = T))
+  dat$id.ordered <- factor(x = dat$ID, levels = c(7,6,5,4,3,2,1)) #order(dat$t2, decreasing = T))
   
   # Use ggplot to plot data from dat object
   ggplot(dat %>% 
-           dplyr::mutate(censored=ifelse(censored==1,"Dead","Censored")) %>%
+           dplyr::mutate(censored=ifelse(censored==1,"Censored","Dead")) %>%
            dplyr::mutate(left_truncated=ifelse(t2<2011,"Left truncated","Included in study")),
          aes(x = id.ordered),group=id.ordered) + 
     # Plot solid line representing non-interval censored time from 0 to t1
[…]

### 2021-01-07 manuscript.tex
update discussion and follow-up plot

@@
 Statistical models that attempt to study the effects of many predictors on survival typically implement feature selection or penalized methods to mitigate the undesirable consequences of overfitting.
-In some cases survival data is also left-truncated which can create an immortal time bias, but penalized survival methods that adjust for left truncation are unavailable in common software.
+In some cases survival data is also left-truncated which can create an immortal time bias, but penalized survival methods that adjust for left truncation are not commonly implemented.
 To address these challenges, we implement a penalized Cox proportional hazards model for left-truncated and right-censored survival data and assess implications of left truncation adjustment on bias and inference.
 We use simulation studies and examples from high-dimensional real-world EHR and genomics data to highlight the pitfalls of failing to account for left truncation in survival modeling.
-The extension of survival modeling to left-truncated data is available in the latest release of \pkg{glmnet}.
 \end{abstract}
@@
 Survival is defined as the time from diagnosis until death, but the study does not begin until 2011 (dashed blue vertical line).
-Patients 2 and 5 (in red) are excluded from the analysis because of death and right censoring, respectively, prior to study eligibility.
-A naive analysis on this data would therefore be biased because it only contains survival time for ``super-survivors'' (those who have survived a long time, such as patients 1 and 3) and the recently-diagnosed.
+Patients 1 and 2 (in red) are excluded from the analysis because of death and right censoring, respectively, prior to study eligibility.
+A naive analysis on this data would therefore be biased because it only contains survival time for ``super-survivors'' (those who have survived a long time, such as patients 3 and 7) and the recently-diagnosed.
 
@@
 The high C-index in the non-adjusted model was artificial because patients diagnosed earlier appeared to survive longer due to immortal time bias.
+To facilitate model interpretation, we recommend that models should be evaluated using both discrimination and calibration based metrics.
 
@@
 
-We conclude with several recommendations.
-When making predictions with survival data is of interest, it is critical to define the risk set appropriately.
+When making predictions with survival data is of interest, it is critical to define the risk set appropriately, as our work demonstrates.
 To do this requires recognizing the underlying process by which individuals are captured in the data in the first place; i.e., when and how individuals came to be included in the dataset of interest.
@@
 While the cohort of lung cancer patients in our real-world use case was restricted to stage IV diagnoses, mitigating this potential selection bias \citep{kehl2020selectionbias}, it can be adjusted for by modeling the association between left truncation time and survival \citep{thiebaut2004choice, chiou2019transformation}.
-Finally, we recommend that models should be evaluated using both discrimination and calibration based metrics.
-
-The implementation of this approach is available in \pkg{glmnet} v.X.X.X on CRAN.
 

### 2021-01-08 supplement.tex
Add Supplement

\documentclass[11pt,final,fleqn]{article}

\usepackage{alltt}

% Basic packages
\usepackage[margin=1in] { geometry }

% Figures
\usepackage[font={bf}]{caption}
\captionsetup{belowskip=10pt,aboveskip=-5pt}
\usepackage{float}
\usepackage[export]{adjustbox}
\renewcommand{\thefigure}{S\arabic{figure}}

% Title
\title{Supplemental Appendix: Penalized regression for left-truncated and right-censored survival data}
\author{Author list}
\date{\today}

\begin{document}
\maketitle

\begin{figure}[h]
\centering
\includegraphics[max size={\textwidth}]{figs/sim_calibration_cumhaz.pdf} 
\caption{Cumulative hazards among non-small cell lung cancer patients in the CGDB in models adjusting for right censoring and left truncation}
\begin{minipage}{\linewidth}
\footnotesize
Notes: The left and right facets plot cumulative hazards for overall survival (i.e., time to death) and followup time (i.e., time to right censoring), respectively. Estimates in the rightmost plot use a ``reverse Kaplan-Meier'' approach whereby the meaning of an event and censoring are flipped. The spline models were fit with one internal knot at the median of the log of followup time. 
\end{minipage}
\end{figure}

\begin{figure}[h]
\centering
\includegraphics[max size={\textwidth}]{figs/sim_int_km_plot_rwd_v_sim.pdf} 
\caption{Kaplan-Meier plots of overall survival from models that adjust for right censoring but not left truncation in the simulated data and CGDB}
\begin{minipage}{\linewidth}
\footnotesize
Notes: The simulated dataset was simulated using an intercept only Weibull model with no predictors. Only non-truncated patients from the simulation were used for estimation. 
\end{minipage}
\end{figure}

\end{document}

### 2021-01-08 R/calibrate_sim.R

@@ -12,7 +12,7 @@ fit_survregs <- function(f, data){
   return(fits)
 }
 
-plot_cumhaz <- function(data){
+plot_cumhaz <- function(data, grp = NULL){
   # Mnual line and colors
   model_names <- unique(data$model)
   n_models <-length(model_names)
@@ -22,11 +22,17 @@ plot_cumhaz <- function(data){
   names(linetypes) <- names(colors) <- c("Kaplan-Meier", parametric_model_names)
   
   # Plot
-  ggplot(data, aes(x = time/365.25, y = est, col = model,
-                   linetype = model)) +
-    geom_line() + xlab("Time") + ylab("Cumulative hazard") +
+  p <- ggplot(data, aes(x = time/365.25, y = est, col = model,
+              linetype = model)) +
+    geom_line() + xlab("Years") + ylab("Cumulative hazard") +
     scale_color_manual(name = "", values = colors) +
-    scale_linetype_manual(name = "", values = linetypes) 
+    scale_linetype_manual(name = "", values = linetypes) +
+    theme(legend.position = "bottom")
+  
+  if (!is.null(grp)) {
+    p <- p + facet_wrap(grp)
+  }
+  return(p)
 }
 
 compare_survregs <- function(fits, km){
@@ -93,6 +99,15 @@ make_survregs <- function(survregfits, coxfit){
   return(fits)
 }
 
+plot_cumhaz_rc_os <- function(cumhaz_rc, cumhaz_os) {
+  pdata <- rbind(
+    as.data.table(cumhaz_rc)[, facet := "Followup"],
+    as.data.table(cumhaz_os)[, facet := "Overall survival"]
+  )
+  pdata[, facet := factor(facet, labels = c("Overall survival", "Followup"))]
+  plot_cumhaz(pdata, grp = "facet")
+}
+
 #' @export
 calibrate_sim <- function(f, data){
   xy <- make_xy(data, f)
@@ -102,18 +117,29 @@ calibrate_sim <- function(f, data){
   }
   
   # Overall survival
-  f_os <- survival::Surv(entry_days_dx, os_days_dx, died) ~ .
-  os_km <- survival::survfit(make_f(f_os, ~1), data) 
-  os_base <- fit_survregs(make_f(f_os, ~1), data)
-  os_cox <- survival::coxph(make_f(f_os, f), data)
[…]

### 2021-01-08 R/run_sim.R

@@ -283,6 +283,25 @@ plot_km <- function(fits){
            linetype = guide_legend(ncol = 1))
 }
 
+plot_km2 <- function(km_sim, km_rwd){
+  surv_df <- rbind(
+    data.table(label = "Simulation",
+               time = km_sim$time,
+               prob = km_sim$surv),
+    data.table(label = "CGDB",
+               time = km_rwd$time/365.25,
+               prob = km_rwd$surv)
+    
+  )
+  ggplot(surv_df,
+         aes(x = time, y = prob, col = label)) + 
+    geom_line() + 
+    scale_color_discrete(name = "") +
+    xlab("Years") + ylab("Survival probability") +
+    theme(legend.position = "bottom") +
+    coord_cartesian(xlim = c(0, 15))
+}
+
 summarize_km <- function(fits, probs = c(.1, .25, .5, .75, .9)){
   n_fits <- length(fits)
   q_mat <- matrix(NA, nrow = n_fits, ncol = length(probs))
@@ -301,7 +320,7 @@ summarize_km <- function(fits, probs = c(.1, .25, .5, .75, .9)){
 }
 
 #' @export
-summarize_simdata <- function(data, save = FALSE, ...) {
+summarize_simdata <- function(data, km_rwd = NULL, save = FALSE, ...) {
   # Descriptive statistics for simulated survival data
   latent_surv_plot <- plot_latent_surv(data) + coord_cartesian(xlim = c(0, 10))
   prop_rc <- 1 - mean(data$dead) 
@@ -314,6 +333,11 @@ summarize_simdata <- function(data, save = FALSE, ...) {
   km_plot <- plot_km(km_fits) + coord_cartesian(xlim = c(0, 10))
   km_quantiles <- summarize_km(km_fits)
   
+  # Compare to RWD
+  if (!is.null(km_rwd)) {
+    km_plot_rwd_v_sim <- plot_km2(km_fits[[3]]$fit, km_rwd)
+  }
+  
   # Return and optionally save
   res <- list(
     latent_surv_plot = latent_surv_plot,
@@ -322,6 +346,7 @@ summarize_simdata <- function(data, save = FALSE, ...) {
     latent_entry_plot = latent_entry_plot,
     stratified_entry_plot = stratified_entry_plot,
     km_plot = km_plot,
+    km_plot_rwd_v_sim = km_plot_rwd_v_sim,
     km_quantiles = km_quantiles
   )
   
[…]

### 2021-01-08 main.R

@@ -114,17 +114,17 @@ rm(test_data)
 # Calibrate simulation ---------------------------------------------------------
 sim_settings <- calibrate_sim(f_small, data = data)
 
-# Cumulative hazard plot comparing different parametric models
-# with Kaplan-Meier estimator
+# Save cumulative hazard plot
 ggsave("figs/sim_calibration_cumhaz.pdf", 
-       sim_settings$os_comparisons$cumhaz_plot,
-       height = 5, width = 7)
+       sim_settings$cumhaz_plot,
+       height = 5, width = 10)
 
 # Example simulated data -------------------------------------------------------
 # No predictors (intercept only model)
 params <- set_params(sim_settings = sim_settings, dist = "weibullPH")
-simdata <- sim_survdata(params = params, n_pats = 5000)
-simdata_summary <- summarize_simdata(simdata, save = TRUE, name = "int")
+simdata <- sim_survdata(params = params, n_pats = 10000)
+simdata_summary <- summarize_simdata(simdata, km_rwd = sim_settings$os_km$rc,  
+                                     save = TRUE, name = "int")
 
 # Now include predictors (and hence create informative censoring when
 # using the Kaplan-Meier estimator). The design matrix X and parameters will 

### 2021-01-08 supplement.tex
Update to Figure S2 in Supplement

@@
 \centering
-\includegraphics[max size={\textwidth}]{figs/sim_int_km_plot_rwd_v_sim.pdf} 
+\includegraphics[max size={\textwidth}]{figs/sim_p11_km_plot_rwd_v_sim.pdf} 
 \caption{Kaplan-Meier plots of overall survival from models that adjust for right censoring but not left truncation in the simulated data and CGDB}
@@
 \footnotesize
-Notes: The simulated dataset was simulated using an intercept only Weibull model with no predictors. Only non-truncated patients from the simulation were used for estimation. 
+Notes: The simulated dataset was simulated using a Weibull model including the predictors described in the main text from the CGDB.
+Only non-truncated patients from the simulation were used for estimation.
 \end{minipage}

### 2021-01-08 R/run_sim.R

@@ -336,6 +336,8 @@ summarize_simdata <- function(data, km_rwd = NULL, save = FALSE, ...) {
   # Compare to RWD
   if (!is.null(km_rwd)) {
     km_plot_rwd_v_sim <- plot_km2(km_fits[[3]]$fit, km_rwd)
+  } else {
+    km_plot_rwd_v_sim <- NULL
   }
   
   # Return and optionally save
@@ -368,9 +370,11 @@ save_simdata <- function(object, name = NULL) {
   ggsave(paste0(fig_path, "km_plot.pdf"), 
          object$km_plot, 
          height = 5, width = 7)
-  ggsave(paste0(fig_path, "km_plot_rwd_v_sim.pdf"), 
-         object$km_plot_rwd_v_sim, 
-         height = 5, width = 7)
+  if (!is.null(object$km_plot_rwd_v_sim)) {
+    ggsave(paste0(fig_path, "km_plot_rwd_v_sim.pdf"), 
+           object$km_plot_rwd_v_sim, 
+           height = 5, width = 7) 
+  }
   write.csv(object$km_quantiles, paste0(tbl_path, "km_quantiles.csv"))
   saveRDS(object$data, paste0(output_path, "data.rds"))
 }

### 2021-01-08 main.R

@@ -123,16 +123,17 @@ ggsave("figs/sim_calibration_cumhaz.pdf",
 # No predictors (intercept only model)
 params <- set_params(sim_settings = sim_settings, dist = "weibullPH")
 simdata <- sim_survdata(params = params, n_pats = 10000)
-simdata_summary <- summarize_simdata(simdata, km_rwd = sim_settings$os_km$rc,  
-                                     save = TRUE, name = "int")
+simdata_summary_int <- summarize_simdata(simdata, km_rwd = sim_settings$os_km$rc,  
+                                        save = TRUE, name = "int")
 
 # Now include predictors (and hence create informative censoring when
 # using the Kaplan-Meier estimator). The design matrix X and parameters will 
 # be used in the simulations that follow
-x_sim <- sim_x(n_pats = 5000, sim_settings, p_bin = 10)
+x_sim <- sim_x(n_pats = 5000, sim_settings, p_bin = 0)
 params <- set_params(x_sim, sim_settings, dist = "weibullPH")
 simdata <- sim_survdata(x_sim, params)
-simdata_summary <- summarize_simdata(simdata, save = TRUE, name = "p10")
+simdata_summary_p11 <- summarize_simdata(simdata, km_rwd = sim_settings$os_km$rc,
+                                         save = TRUE, name = "p11")
 
 # Run simulation for unpenalized Cox model -------------------------------------
 sim_coxph_p21 <- xfun::cache_rds({
@@ -399,6 +400,8 @@ p_cal_coxlasso <- ggarrange(p_cal_coxlasso_small, p_cal_coxlasso_large,
 ggsave("figs/calibration_coxlasso.pdf", p_cal_coxlasso, width = 7, height = 9)
 
 # Save text statistics ---------------------------------------------------------
+txt$propTruncatedSim <- formatC(simdata_summary_p11$prop_truncated, 
+                                format = "f", digits = 3)
 txt$nPatients <- formatC(nrow(data), format = "d", big.mark = ",")
 txt$nDeaths <- formatC(n_deaths, format = "d", big.mark = ",")
 txt$maxP <- formatC(as.integer(max_p), format = "d", big.mark = ",")

### 2021-01-08 main.R
Results update

@@ -401,7 +401,7 @@ ggsave("figs/calibration_coxlasso.pdf", p_cal_coxlasso, width = 7, height = 9)
 
 # Save text statistics ---------------------------------------------------------
 txt$propTruncatedSim <- formatC(simdata_summary_p11$prop_truncated, 
-                                format = "f", digits = 3)
+                                format = "f", digits = 2)
 txt$nPatients <- formatC(nrow(data), format = "d", big.mark = ",")
 txt$nDeaths <- formatC(n_deaths, format = "d", big.mark = ",")
 txt$maxP <- formatC(as.integer(max_p), format = "d", big.mark = ",")

### 2021-01-08 manuscript.tex
update reference style

@@
 \usepackage{natbib}
-\bibpunct{(}{)}{;}{a}{}{,}
-\bibliographystyle{apa}
+\setcitestyle{super}
+\bibliographystyle{WileyNJD-AMA}
+\setcitestyle{open={},close={},comma,sort&compress}
 \renewcommand{\bibname}{References}
+\renewcommand\bibnumfmt[1]{{#1}.} % to format bibliography as 1. and 2. etc
 
@@
 The amount of person-level and disease-specific detail captured by these technologies offers an exciting new opportunity to understand patient history and prognosis by combining these molecular data with macro-scale clinical data on patient characteristics, treatment, and outcomes.
-Specifically, these data may offer insights into risk factors affecting patient-related outcomes leading to interest in developing prognostic tools to improve clinical decision-making and patient care \citep{gui2005penalized, wishart2010predict, ow2016big, yousefi2017predicting}.
+Specifically, these data may offer insights into risk factors affecting patient-related outcomes leading to interest in developing prognostic tools to improve clinical decision-making and patient care \cite{gui2005penalized, wishart2010predict, ow2016big, yousefi2017predicting}.
 Given the hundreds or even thousands of molecular, patho-physiological, and clinical parameters that may be readily available for each patient, developing an appropriate statistical model that is robust to overfitting is critical to ensure good model performance and generalizability.
@@
 
-In a high-dimensional setting, computationally efficient regularization and shrinkage techniques have been developed \citep{tibshirani1997lasso, friedman2010regularization, simon2011regularization} for the Cox model.
+In a high-dimensional setting, computationally efficient regularization and shrinkage techniques have been developed \cite{tibshirani1997lasso, friedman2010regularization, simon2011regularization} for the Cox model.
 Similar to ordinary least squares (OLS) regression, these penalized models estimate regression coefficients by minimizing the sum of squared errors, but place a constraint in the equation to minimize the sum of coefficients.
 This constraint (controlled by a parameter, $\lambda$) penalizes a high number of variables in the model and shrinks the coefficients of less influential features towards zero - acting, in cases when some coefficients are shrunk fully to zero, as automatic feature selection.
-Common penalized approaches include lasso \citep{tibshirani1996regression}, ridge \citep{hoerl1970ridgeA, hoerl1970ridgeB}, and elastic net \citep{zou2005regularization}, with extensions such as the adaptive lasso \citep{zou2006adaptive}.
+Common penalized approaches include lasso \cite{tibshirani1996regression}, ridge \cite{hoerl1970ridgeA, hoerl1970ridgeB}, and elastic net \cite{zou2005regularization}, with extensions such as the adaptive lasso \cite{zou2006adaptive}.
 
@@
 In other words, the data are left-truncated because patients are only included if they survive beyond some initial milestone that marks entry into the observed data.
-An analysis of the observed patients is subject to an immortal time bias because observed patients cannot die prior to entering the study cohort \citep{levesque2010problem, giobbie2013challenges}.
+An analysis of the observed patients is subject to an immortal time bias because observed patients cannot die prior to entering the study cohort \cite{levesque2010problem, giobbie2013challenges}.
 An example in medicine is a dataset describing patients who receive a specific diagnostic or genetic test; patients who die before having the chance to be tested are excluded from the data, and consequently their survival times do not contribute to overall survival estimates.
 Traditional survival models are inappropriate in this case and predicted survival probabilities are overestimated because of the exclusion of short survival times.
-Such bias has been observed in large oncology cohorts marked by a milestone entry, including patients who receive genomic tests \citep{kehl2020selectionbias} and specialized treatments \citep{newman2020immortaltime}.
+Such bias has been observed in large oncology cohorts marked by a milestone entry, including patients who receive genomic tests \cite{kehl2020selectionbias} and specialized treatments \cite{newman2020immortaltime}.
 These databases are increasingly used to understand patient survival to support clinical decision-making, motivating the need to identify and address this survivor selection bias in practice.
 
-Left truncation is a well known feature of medical data and there are methods to adjust for it \citep{kalbfleisch2011statistical}.
-The Kaplan-Meier estimator of the survival function requires almost no modifications as it is only necessary to change the risk set so that individuals are only at risk of an event after study entry \citep{tsai1987note}.
+Left truncation is a well known feature of medical data and there are methods to adjust for it \cite{kalbfleisch2011statistical}.
+The Kaplan-Meier estimator of the survival function requires almost no modifications as it is only necessary to change the risk set so that individuals are only at risk of an event after study entry \cite{tsai1987note}.
 Similarly, with left truncation, the likelihood function in a parametric model can be adjusted so that it is conditional on survival to the entry time.
@@
 
-\citet{cox1972regression} proposed a partial likelihood for $\beta$ without involving the baseline hazard $h_0(t)$
+Cox\cite{cox1972regression} proposed a partial likelihood for $\beta$ without involving the baseline hazard $h_0(t)$
 
@@
 \subsection{Penalized regression with left-truncated and right-censored data}
-Penalized regression methods such as $l_1$ (lasso) \citep{tibshirani1996regression} and $l_2$ (ridge) \citep{tikhonov1963ridge} regression, elastic net regression \citep{zou2005regularization}, and SCAD (Smoothly Clipped Absolute Deviation) \citep{xie2009scad} have been developed to overcome the limitation of fitting a model to a large number of predictors.
+Penalized regression methods such as $l_1$ (lasso) \cite{tibshirani1996regression} and $l_2$ (ridge) \cite{tikhonov1963ridge} regression, elastic net regression \cite{zou2005regularization}, and SCAD (Smoothly Clipped Absolute Deviation) \cite{xie2009scad} have been developed to overcome the limitation of fitting a model to a large number of predictors.
 In a high dimensional feature space, where the number of predictors is larger than the sample size (i.e., $p$ > $n$), a model may exhibit low bias but suffer from poor prediction accuracy and limited generalizability, a consequence of overfitting the data.
@@
 
-The extension of penalized regression methods to survival data have been extensively described and applied, for example in microarray gene expression data \citep{gui2005penalized}, transcriptomes \citep{wu2011penalized}, and injury scale codes \citep{mittal2013penalized}.
+The extension of penalized regression methods to survival data have been extensively described and applied, for example in microarray gene expression data \cite{gui2005penalized}, transcriptomes \cite{wu2011penalized}, and injury scale codes \cite{mittal2013penalized}.
 Like linear regression, the performance and fit of a Cox proportional hazards model is known to deteriorate with large $p$, providing an advantage to penalized approaches in high-dimensional, time-to-event feature sets.
 
-\citet{simon2011regularization} introduced a fast, pathwise algorithm to fit a regularized Cox proportional hazards model solving for
+Simon et. al\cite{simon2011regularization} introduced a fast, pathwise algorithm to fit a regularized Cox proportional hazards model solving for
 
@@
 where $\beta$ is the vector of regression coefficients and is maximized over \autoref{eqn:simon} subject to a constraint defined by the penalty $\lambda \cdot P_{\alpha}\beta$.
-As described elsewhere \citep{simon2011regularization}, the penalty can take the form of the $l_1$ (lasso), $l_2$ (ridge), or elastic net penalty,  and the algorithm fits over a path of $\lambda$ values via cyclical coordinate descent.
+As described elsewhere \cite{simon2011regularization}, the penalty can take the form of the $l_1$ (lasso), $l_2$ (ridge), or elastic net penalty,  and the algorithm fits over a path of $\lambda$ values via cyclical coordinate descent.
 Instead of solving a least squares problem at each iteration, the algorithm solves a penalized reweighted least squares problem, in which observation weights $w$ are defined in the coordinate descent least squares minimization step.
 The optimal $\lambda$ can be identified through cross-validation.
-Tied survival times are handled via the Breslow approximation \citep{breslow1972} and details provided in \citep{simon2011regularization}.
+Tied survival times are handled via the Breslow approximation \cite{breslow1972} and details provided in \cite{simon2011regularization}.
 
@@
 \subsection{Results}
-We used calibration curves \citep{harrell2015regression} to compare models with and without a left truncation adjustment.
+We used calibration curves \cite{harrell2015regression} to compare models with and without a left truncation adjustment.
 Each point in the plot consists of simulated patients in a given decile of predicted survival at a given time point.
@@
 Predictions tend to be more accurate for patients with predicted survival close to and below $0.5$, but slightly underestimated for patients with higher survival probabilities.
-Probability rescaling may be effective in such cases, for example with the use of Platt scaling \citep{platt1999scaling} or isotonic regression \citep{niculescu2005isotonic}.
+Probability rescaling may be effective in such cases, for example with the use of Platt scaling \cite{platt1999scaling} or isotonic regression \cite{niculescu2005isotonic}.
 
-A common metric used to evaluate survival models is the C-index \citep{harrell1996multivariable}, which provides a measure of how well a model will discriminate prognosis between patients.
+A common metric used to evaluate survival models is the C-index \cite{harrell1996multivariable}, which provides a measure of how well a model will discriminate prognosis between patients.
 We computed the C-index using predictions from the Cox model with lasso penalty on high dimensional data.
@@
 Survival was predicted from diagnosis.
-For an evaluation of meaningful prognostic clinical and genomic factors in NSCLC, see for example \citet{lai2020nsclc}.
+For an evaluation of meaningful prognostic clinical and genomic factors in NSCLC, see for example Lai and colleagues\cite{lai2020nsclc}.
 
@@
 As the data come from both historic and current clinico-genomic testing products, the set of tested genes has changed over time.
-To resolve the resultant missing data problem, we constrained the mutation data to biomarkers with less than 30\% missingness and imputed the remaining mutation data using k nearest neighbors \citep{troyanskaya_missing_2001}, with k = 20.
+To resolve the resultant missing data problem, we constrained the mutation data to biomarkers with less than 30\% missingness and imputed the remaining mutation data using k nearest neighbors \cite{troyanskaya_missing_2001}, with k = 20.
 Biomarkers with zero variance or that were present in less than 0.1\% of patients were also excluded.
@@
 One of the mutation features that had the highest importance was the presence of short variants in the gene KEAP1.
-This finding is supported by similar reports of risk-conferring status in an independent lung cancer cohort \citep{shen_harnessing_2019}.
+This finding is supported by similar reports of risk-conferring status in an independent lung cancer cohort \cite{shen_harnessing_2019}.
 Conversely, variants in EGFR and ALK were found to be protective in the CGDB data, putatively because mutations in these genes qualify patients to receive targeted therapies and therefore benefit from improved response and longer survival.
@@
 The latter may occur if left truncation time is correlated with survival time, for example, if genomic testing is ordered selectively for patients with worse prognosis.
-This was uncovered in a cohort of multi-stage cancer patients who received genomic testing \citep{kehl2020selectionbias} and can affect inference even in the presence of a left truncation adjustment.
-While the cohort of lung cancer patients in our real-world use case was restricted to stage IV diagnoses, mitigating this potential selection bias \citep{kehl2020selectionbias}, it can be adjusted for by modeling the association between left truncation time and survival \citep{thiebaut2004choice, chiou2019transformation}.
+This was uncovered in a cohort of multi-stage cancer patients who received genomic testing \cite{kehl2020selectionbias} and can affect inference even in the presence of a left truncation adjustment.
+While the cohort of lung cancer patients in our real-world use case was restricted to stage IV diagnoses, mitigating this potential selection bias \cite{kehl2020selectionbias}, it can be adjusted for by modeling the association between left truncation time and survival \cite{thiebaut2004choice, chiou2019transformation}.
 

### 2021-01-08 manuscript.tex
add CGDB citation

@@
 To address these challenges, we implement a penalized Cox proportional hazards model for left-truncated and right-censored survival data and assess implications of left truncation adjustment on bias and inference.
-We use simulation studies and examples from high-dimensional real-world EHR and genomics data to highlight the pitfalls of failing to account for left truncation in survival modeling.
+We use simulation studies and a high-dimensional, real-world clinico-genomic database (CGDB) to highlight the pitfalls of failing to account for left truncation in survival modeling.
 \end{abstract}
@@
 \subsection{Real-world use case}
-We consider the Clinico-Genomic Database (CGDB) offered jointly by Flatiron Health and Foundation Medicine.
-The CGDB is an oncology research platform that combines real-world, patient-level clinical data and outcomes with patient-level genomic data to provide a comprehensive patient and tumor profile.
-Patients in the CGDB have received at least one Foundation Medicine tumor genomic test as well as have had their electronic health records captured by Flatiron Health.
-Together, these data sources represent thousands of potential predictors per patient, ranging from demographic, laboratory, and treatment information to the specific mutations detected on each biomarker included in a Foundation Medicine baitset.
+We consider the Clinico-Genomic Database (CGDB) offered jointly by Flatiron Health and Foundation Medicine \cite{singal2017cgdb}.
+The CGDB is a US-based oncology research platform that combines real-world, patient-level clinical data and outcomes with patient-level genomic data to provide a comprehensive patient and tumor profile.
+The de-identified data originated from approximately 280 US cancer clinics (~800 sites of care).
+Patients in the CGDB have received at least one Foundation Medicine comprehensive genomic profiling test as well as have had their electronic health records captured by Flatiron Health.
+The database comprises patient-level structured and unstructured data, curated via technology-enabled abstraction, linked to Foundation Medicine genomic data by de-identified, deterministic matching.
+The data are de-identified and subject to obligations to prevent re-identification and protect patient confidentiality.
+Altogether, the CGDB represents thousands of potential predictors per patient, ranging from demographic, laboratory, and treatment information to the specific mutations detected on each biomarker included in a Foundation Medicine baitset.
 

### 2021-01-08 manuscript.tex
add Fig S1 and S2

@@
 The Nelson-Aalen estimator of the cumulative hazard of overall survival was generally monotonically decreasing, which suggested that a Weibull distribution could adequately capture the baseline hazard.
-This was further supported by the similarity in fit between a spline-based survival model with one internal knot at the median of uncensored survival time and the Weibull model.
+This was further supported by the similarity in fit between a spline-based survival model with one internal knot at the median of uncensored survival time and the Weibull model (Figure S1).
 As such, an intercept only Weibull model was used to estimate $a$ and $\alpha$ in the simulation.
-Similarly, a Weibull distribution was used to model time to right censoring (by reversing the censoring indicator in the model for overall survival.
+Similarly, a Weibull distribution was used to model time to right censoring by reversing the censoring indicator in the model for overall survival.
 Both the model for overall survival and right censoring adjusted for left truncation.
@@
 The probability of a positive entry time was set equal to $0.2$ and the lognormal distribution for positive entry times was derived using a right skewed distribution with a median value of $1$ and a mean value of $1.6$, both measured in years.
-We found that this parameterization generated median survival times in the simulation that were similar to median survival in Kaplan-Meier estimates that did not adjust for left truncation in the CGDB.
+We found that this parameterization generated median survival times in the simulation that were similar to median survival in Kaplan-Meier estimates that did not adjust for left truncation in the CGDB (Figure S2).
 

### 2021-01-11 manuscript.tex
add acknowledgments, data availability statement

@@
 
-\pdfbookmark[1]{References}{References}
+\section*{Acknowledgments} 
+We thank T. Hastie and K. Tay for their work in implementing left truncation-adjusted Cox models in \pkg{glmnet}.
+
+\section*{Data availability statement}
+The data that support the findings of this study have been originated by Flatiron Health, Inc.
+and Foundation Medicine, Inc.
+These de-identified data may be made available upon request, and are subject to a license agreement with Flatiron Health and Foundation Medicine; interested researchers should contact <cgdb-fmi@flatiron.com> to determine licensing terms.
+
+Code used to produce the analysis is publicly available at https://github.com/phcanalytics/coxnet-ltrc.
+
+%\section*{References} 
+%\pdfbookmark[3]{References}{References}
 \bibliography{references}

### 2021-01-12 manuscript.tex
add glmnet v.4.1 citation

@@
 Furthermore, we show that use of inappropriate metrics for model evaluation can result in overly optimistic conclusions about the performance of predictive models.
-For implementation of the proposed method, we use the left truncation extension to survival modeling introduced in v.X.X of the \textsf{R} package \pkg{glmnet} (CITE LATEST PACKAGE).
+For implementation of the proposed method, we use the left truncation extension to survival modeling introduced in v.4.1 of the \textsf{R} package \pkg{glmnet} \cite{glmnet}.
 
@@
 This implementation is available in the release of \pkg{glmnet} v.
-X.X (CITE LATEST PACKAGE).
+4.1 \cite{glmnet}.
 

### 2021-01-12 manuscript.tex
Consistent citation of glmnet

@@
 Furthermore, we show that use of inappropriate metrics for model evaluation can result in overly optimistic conclusions about the performance of predictive models.
-For implementation of the proposed method, we use the left truncation extension to survival modeling introduced in v.4.1 of the \textsf{R} package \pkg{glmnet} \cite{glmnet}.
+For implementation of the proposed method, we use the left truncation extension to survival modeling introduced in version 4.1 of the \textsf{R} package \pkg{glmnet} \cite{glmnet}.
 
@@
 An efficient implementation of this method is non-trivial, as the usual updating of risk set components cannot be used when left truncation is present.
-This implementation is available in the release of \pkg{glmnet} v.
-4.1 \cite{glmnet}.
+This implementation is available in the release of \pkg{glmnet} version 4.1 \cite{glmnet}.
 

### 2021-01-27 manuscript.tex
Add clickable link to GitHub repo in manuscript
Flatiron manuscript edits

@@
 \begin{abstract}
-    High-dimensional data are becoming increasingly common in the medical field as large volumes of patient information are collected and processed by high-thoroughput screening, electronic health records (EHRs), and comprehensive genomic testing.
+    High-dimensional data are becoming increasingly common in the medical field as large volumes of patient information are collected and processed by high-throughput screening, electronic health records (EHRs), and comprehensive genomic testing.
 Statistical models that attempt to study the effects of many predictors on survival typically implement feature selection or penalized methods to mitigate the undesirable consequences of overfitting.
@@
 Left truncation is a well known feature of medical data and there are methods to adjust for it \cite{kalbfleisch2011statistical}.
-The Kaplan-Meier estimator of the survival function requires almost no modifications as it is only necessary to change the risk set so that individuals are only at risk of an event after study entry \cite{tsai1987note}.
+Given quasi-independence between left truncation time and study entry\cite{tsai1987note}, the Kaplan-Meier estimator of the survival function requires almost no modifications as it is only necessary to change the risk set so that individuals are only at risk of an event after study entry \cite{tsai1987note}.
 Similarly, with left truncation, the likelihood function in a parametric model can be adjusted so that it is conditional on survival to the entry time.
@@
 Here, we implement a penalized Cox proportional hazards model for left-truncated and right-censored (LTRC) survival data and critically assess scenarios in which bias arises from and is exacerbated by left truncation.
-We apply the method to simulated data and a real-world use case from the Flatiron Health and Foundation Medicine Clinico-Genomic Database (CGDB).
+We apply the method to simulated data and a real-world use case from a de-identified clinico-genomic database (CGDB).
 Our results highlight the importance of adjusting for left truncation for both inference and prediction.
@@
 
-The remainder of this paper is organized as follows: In \autoref{sec:methods}, we introduce the penalized regression and left truncation framework and present the efficient method that incorporates both.
+The remainder of this paper is organized as follows: In \autoref{sec:methods}, we introduce the penalized regression and left truncation framework and present the method that incorporates both.
 \autoref{sec:simulation} presents the simulation results and \autoref{sec:rwd} considers a real-world dataset that combines electronic health records and genomic testing information.
@@
 
-Fortunately, the partial likelihood can be easily modified to make $R_i$ the set of individuals who are at risk (have entered the study and are not censored or failed) at that time.
+Fortunately, the partial likelihood can be easily modified to make $R_i$ the set of individuals who are at risk (have entered the study and are not censored or failed) at that time assuming quasi-independence between left truncation and failure time \cite{tsai1987note}.
 Let individual $i$ have survival data of the form $(v_i, y_i, e_i, x_i)$, where individual $i$ enters the study at time $v_i$.
@@
 exceeded their left truncation time).
-Here, event times $t_1 < t_2 < ...
-< t_m$ are ordered relative to the earliest chronological event time observed in the patient cohort (as shown in \autoref{fig:followup}), rather than assuming that all patients are at risk starting at the same referent time $0$.
+For example, the at-risk time for patients in Figure 1 is counted only after the vertical blue line marking study entry.
 
@@
 where the left-truncated risk set $R_{i\ast} = \{ j : y_j \geq t_i > v_j \}$; that is, subjects are only considered at risk after their left truncation times.
-An efficient implementation of this method is non-trivial, as the usual updating of risk set components cannot be used when left truncation is present.
 This implementation is available in the release of \pkg{glmnet} version 4.1 \cite{glmnet}.
@@
 \subsection{Real-world use case}
-We consider the Clinico-Genomic Database (CGDB) offered jointly by Flatiron Health and Foundation Medicine \cite{singal2017cgdb}.
-The CGDB is a US-based oncology research platform that combines real-world, patient-level clinical data and outcomes with patient-level genomic data to provide a comprehensive patient and tumor profile.
+We consider the Clinico-Genomic Database offered jointly by Flatiron Health and Foundation Medicine (FH-FMI CGDB) \cite{singal2017cgdb}.
+The CGDB is a US-based de-identified oncology database that combines real-world, patient-level clinical data and outcomes with patient-level genomic data to provide a comprehensive patient and tumor profile.
 The de-identified data originated from approximately 280 US cancer clinics (~800 sites of care).
 Patients in the CGDB have received at least one Foundation Medicine comprehensive genomic profiling test as well as have had their electronic health records captured by Flatiron Health.
-The database comprises patient-level structured and unstructured data, curated via technology-enabled abstraction, linked to Foundation Medicine genomic data by de-identified, deterministic matching.
+The retrospective, longitudinal EHR-derived database comprises patient-level structured and unstructured data, curated via technology-enabled abstraction, and were linked to Foundation Medicine genomic data by de-identified, deterministic matching.
+Genomic alterations were identified via comprehensive genomic profiling (CGP) of >300 cancer-related genes on FMI's next-generation sequencing (NGS) based FoundationOne panel \cite{birnbaum2020modelassisted, ma2020fmi}.
+To date, over 400,000 samples have been sequenced from patients across the US.
 The data are de-identified and subject to obligations to prevent re-identification and protect patient confidentiality.
 Altogether, the CGDB represents thousands of potential predictors per patient, ranging from demographic, laboratory, and treatment information to the specific mutations detected on each biomarker included in a Foundation Medicine baitset.
+Institutional Review Board approval of the study protocol was obtained prior to study conduct, and included a waiver of informed consent.
 
-We take a subset of $\nPatients$ stage IV, non-small cell lung cancer (NSCLC) patients in the CGDB as a use case for this study.
+We take a subset of $\nPatients$ patients with stage IV, non-small cell lung cancer (NSCLC) in the CGDB as a use case for this study.
 The outcome of interest is survival time predicted from the date of stage IV diagnosis.
@@
 In other words, the bias in the coefficients is based only on the indirect correlation between the predictors and left truncation induced by (i) the correlation between the predictors and survival and (ii) the correlation between survival and the probability a patient is left-truncated.
-As we will see in \autoref{sec:rwd}, the impact of left truncation adjustment on the estimated values of the coefficients (and by extenstion the C-index) can be considerably larger if there is a more direct correlation.
+As we will see in \autoref{sec:rwd}, the impact of left truncation adjustment on the estimated values of the coefficients (and by extension the C-index) can be considerably larger if there is a more direct correlation.
 
-\section{Real-world data application: Non-small cell lung cancer patients in the Clinico-Genomic Database (CGDB)} \label{sec:rwd}
-We fit models using $\nPatients$ stage IV, NSCLC patients.
+\section{Real-world data application: Patients with non-small cell lung cancer in the Clinico-Genomic Database (CGDB)} \label{sec:rwd}
+We fit models using $\nPatients$ patients with stage IV, NSCLC.
 75\%  of patients were assigned to a training set and the remaining 25\% were assigned to a test set.
@@
 \section{Discussion} \label{sec:discussion}
-In this paper, we implemented an efficient approach for estimating penalized Cox proportional hazards model with LTRC survival data and compared it to penalized models designed for survival data that is right-censored but not left-truncated.
+In this paper, we implemented an approach for estimating penalized Cox proportional hazards model with LTRC survival data and compared it to penalized models designed for survival data that is right-censored but not left-truncated.
 Using simulation studies and examples from real-world EHR and genomics data, we showed that there is a need for approaches that can both adjust for left truncation and model high dimensional data.
@@
 
-Code used to produce the analysis is publicly available at https://github.com/phcanalytics/coxnet-ltrc.
+Code used to produce the analysis is publicly available at \url{https://github.com/phcanalytics/coxnet-ltrc}.
 

### 2021-01-28 R/plot_patient_followup.R
update to reflect milestone entry

@@ -5,12 +5,14 @@
 #' @export
 plot_patient_followup <- function(){
   dat <- data.frame(ID = c(1:7), event = c(0, 1, 1, 1, 0, 0, 0),  # c(7, 1, 3, 4, 2, 6, 5)
-                    t1 = c(1991, 2008, 1993, 2007, 2012, 2014, 1989),
+                    t1 = c(1991, 2008, 1993, 2007, 2010, 2012, 1989),
                     t2 = c(1998, 2009, 2013, 2017, 2019, 2019, 2019), 
+                    t3 = c(NA, NA, 2011, 2011, 2014, 2014, 2015),
+                    t4 = c(1998, 2009, 2011, 2011, 2014, 2014, 2015),
                     #t1 = c(1989, 1991, 1993, 2007, 2008, 2012, 2014), 
-                   # t2 = c(2019, 1998, 2013, 2017, 2009, 2019, 2019), 
-                   censored = c(0, 1, 0, 0, 1, 1, 1))
-                   # censored = c(1, 0, 0, 0, 1, 1, 1)) #, .Names = c("ID", "Death", "t1", "t2", "Censored")) #, class = "data.frame", row.names = c(NA, -6L))
+                    # t2 = c(2019, 1998, 2013, 2017, 2009, 2019, 2019), 
+                    censored = c(0, 1, 0, 0, 1, 1, 1))
+  # censored = c(1, 0, 0, 0, 1, 1, 1)) #, .Names = c("ID", "Death", "t1", "t2", "Censored")) #, class = "data.frame", row.names = c(NA, -6L))
   
   # Create event variable
   #dat$Death <- with(dat, ifelse(Death, "Dead", "Censored"))
@@ -25,9 +27,11 @@ plot_patient_followup <- function(){
            dplyr::mutate(left_truncated=ifelse(t2<2011,"Left truncated","Included in study")),
          aes(x = id.ordered),group=id.ordered) + 
     # Plot solid line representing non-interval censored time from 0 to t1
-    geom_linerange(aes(ymin = t1, ymax = t2, col=left_truncated)) + 
+    geom_linerange(aes(ymin = t1, ymax = t4, col=left_truncated), linetype="dashed") + 
+    geom_linerange(aes(ymin = t4, ymax = t2, col=left_truncated)) + 
     geom_point(aes(id.ordered, t2,shape=factor(censored),col=left_truncated)) + 
[…]

### 2021-01-28 manuscript.tex
update LT section and patient follow-up

@@
 In this sense, all individuals enter the risk set at the same time 0 and leave only at death or censoring.
-However, when survival times are left-truncated, individuals are truly at risk prior to entering the study (e.g.
+However, when survival times are left-truncated, individuals are truly at risk prior to entering the study cohort (e.g.
 prior to being observed for follow-up).
@@
 \caption{Left-truncated and right-censored patient follow-up in a hypothetical study cohort, ordered chronologically by event time.
-Patients become eligible to enter the study after diagnosis (closed circle) and are followed until death (open circle) or censoring (cross).
-However, patients who die before the start of the study (dashed blue line) are left-truncated (in red), and only those who have survived until the study start (in black) are observed.}
+Patients who receive a diagnosis (closed circle) become eligible to enter the cohort after reaching a milestone (black triangle), for example a genomic test.
+Patients are followed until death (open circle) or censoring (cross).
+However, patients who die before reaching the milestone are left-truncated (in red), and only those who have survived until eligibility (in black) are observed.
+Left truncation time, or the time between diagnosis and cohort entry, is shown with a dashed line.}
 \label{fig:followup}
@@
 
-Survival is defined as the time from diagnosis until death, but the study does not begin until 2011 (dashed blue vertical line).
+Survival is defined as the time from diagnosis until death, but patients are not eligible to enter the cohort until a milestone is reached (black triangle).
 Patients 1 and 2 (in red) are excluded from the analysis because of death and right censoring, respectively, prior to study eligibility.
@@
 The left-truncated risk set $R_{i\ast} = \{ j : y_j \geq t_i > v_j \}$.
-In other words, subjects are only considered at risk in the analysis after they have been observed to be at risk (i.e.
-exceeded their left truncation time).
-For example, the at-risk time for patients in Figure 1 is counted only after the vertical blue line marking study entry.
+In other words, subjects are only considered at risk at a given event time after they have been observed to be at risk (i.e.
+exceeded their left truncation time, shown with a solid black line in \autoref{fig:followup} ).
+For example, at the time of death of Patient 3 in \autoref{fig:followup} , only Patient 4 has reached the milestone for study entry and is at risk.
+Patients 5, 6, and 7 reach their milestones at later time points.
 
@@
 We generated $n$ patients and $p$ predictors to construct a $n \times p$ matrix, $X$.
-The first 11 predictors were from the CGDB and the remaining $\tilde{p}$ variables were simulated binary variables.
+The first 11 predictors were from the CGDB and the remaining $\tilde{p}$ variables were simulated binary variables to represent a comprehensive set of CGDB binary alterations.
 The CGDB variables were the same $11$ variables used in the small $p$ model in \autoref{sec:rwd} and generated by randomly sampling rows of the original data with replacement.

### 2021-02-02 manuscript.tex
new round of Flatiron edits, authors & affils

@@
 % basic packages
+\usepackage{authblk}
 \usepackage[T1]{fontenc}
@@
 \title{Penalized regression for left-truncated and right-censored survival data}
-\author{Author list}
+
+%\author[1]{Sarah McGough\footnote{Contributed equally.}}
+\newcommand*\samethanks[1][\value{footnote}]{\footnotemark[#1]}
+
+\author[1]{Sarah F. McGough\thanks{Contributed equally.}}
+\author[1]{Devin Incerti\samethanks}
+\author[1]{Svetlana Lyalina}
+\author[1]{Ryan Copping}
+\author[2, 3]{Balasubramanian Narasimhan}
+\author[2, 3]{Robert Tibshirani}
+\affil[1]{Genentech, Inc, South San Francisco, CA, USA}
+\affil[2]{Department of Statistics, Stanford University, Stanford, CA, USA}
+\affil[3]{Department of Biomedical Data Sciences, Stanford University, Stanford, CA, USA}
+
+%\renewcommand\Authands{ , }
+
 \date{\today}
@@
 Statistical models that attempt to study the effects of many predictors on survival typically implement feature selection or penalized methods to mitigate the undesirable consequences of overfitting.
-In some cases survival data is also left-truncated which can create an immortal time bias, but penalized survival methods that adjust for left truncation are not commonly implemented.
-To address these challenges, we implement a penalized Cox proportional hazards model for left-truncated and right-censored survival data and assess implications of left truncation adjustment on bias and inference.
+In some cases survival data is also left-truncated which can give rise to an immortal time bias, but penalized survival methods that adjust for left truncation are not commonly implemented.
+To address these challenges, we implement a penalized Cox proportional hazards model for left-truncated and right-censored survival data and assess implications of left truncation adjustment on bias and interpretation.
 We use simulation studies and a high-dimensional, real-world clinico-genomic database (CGDB) to highlight the pitfalls of failing to account for left truncation in survival modeling.
@@
 An example in medicine is a dataset describing patients who receive a specific diagnostic or genetic test; patients who die before having the chance to be tested are excluded from the data, and consequently their survival times do not contribute to overall survival estimates.
-Traditional survival models are inappropriate in this case and predicted survival probabilities are overestimated because of the exclusion of short survival times.
+Traditional survival models are inappropriate in this case and predicted survival probabilities are overestimated because of the exclusion of short survival times, biasing the study sample and consequently the estimation of the baseline risk and regression coefficients.
 Such bias has been observed in large oncology cohorts marked by a milestone entry, including patients who receive genomic tests \cite{kehl2020selectionbias} and specialized treatments \cite{newman2020immortaltime}.
@@
 Left truncation is a well known feature of medical data and there are methods to adjust for it \cite{kalbfleisch2011statistical}.
-Given quasi-independence between left truncation time and study entry\cite{tsai1987note}, the Kaplan-Meier estimator of the survival function requires almost no modifications as it is only necessary to change the risk set so that individuals are only at risk of an event after study entry \cite{tsai1987note}.
+The Kaplan-Meier estimator of the survival function requires almost no modifications as it is only necessary to change the risk set so that individuals are only at risk of an event after study entry \cite{tsai1987note}.
 Similarly, with left truncation, the likelihood function in a parametric model can be adjusted so that it is conditional on survival to the entry time.
@@
 We apply the method to simulated data and a real-world use case from a de-identified clinico-genomic database (CGDB).
-Our results highlight the importance of adjusting for left truncation for both inference and prediction.
+Our results highlight the importance of adjusting for left truncation for both model interpretation and prediction.
 Furthermore, we show that use of inappropriate metrics for model evaluation can result in overly optimistic conclusions about the performance of predictive models.
@@
 
-Fortunately, the partial likelihood can be easily modified to make $R_i$ the set of individuals who are at risk (have entered the study and are not censored or failed) at that time assuming quasi-independence between left truncation and failure time \cite{tsai1987note}.
+Fortunately, the partial likelihood can be easily modified to make $R_i$ the set of individuals who are at risk (have entered the study and are not censored or failed) at that time.
 Let individual $i$ have survival data of the form $(v_i, y_i, e_i, x_i)$, where individual $i$ enters the study at time $v_i$.
@@
 Penalized regression methods such as $l_1$ (lasso) \cite{tibshirani1996regression} and $l_2$ (ridge) \cite{tikhonov1963ridge} regression, elastic net regression \cite{zou2005regularization}, and SCAD (Smoothly Clipped Absolute Deviation) \cite{xie2009scad} have been developed to overcome the limitation of fitting a model to a large number of predictors.
-In a high dimensional feature space, where the number of predictors is larger than the sample size (i.e., $p$ > $n$), a model may exhibit low bias but suffer from poor prediction accuracy and limited generalizability, a consequence of overfitting the data.
+In a high-dimensional feature space, where the number of predictors $p$ is large relative to the sample size $n$ , a model may exhibit low bias but suffer from poor prediction accuracy and limited generalizability, a consequence of overfitting the data.
 Penalized approaches, which add a regularization penalty to constrain the size of the $\beta$, reduce the complexity of the model and mitigate the negative effects of overfitting.
+Note that in high-dimensional settings where the number of predictors is larger than the sample size (i.e., $p$ > $n$), a standard regression cannot be fit at all.
 
@@
 Predictions tend to be more accurate for patients with predicted survival close to and below $0.5$, but slightly underestimated for patients with higher survival probabilities.
-Probability rescaling may be effective in such cases, for example with the use of Platt scaling \cite{platt1999scaling} or isotonic regression \cite{niculescu2005isotonic}.
+Probability rescaling has been effective in such cases, for example with the use of Platt scaling \cite{platt1999scaling} or isotonic regression \cite{niculescu2005isotonic}, though implementation is less straightforward in a survival context \cite{goldstein2021xcal}.
 
 A common metric used to evaluate survival models is the C-index \cite{harrell1996multivariable}, which provides a measure of how well a model will discriminate prognosis between patients.
-We computed the C-index using predictions from the Cox model with lasso penalty on high dimensional data.
+We computed the C-index using predictions from the Cox model with lasso penalty on high-dimensional data.
 The results are presented in \autoref{tbl:sim-model-discrimination} and suggest that care should be taken when using the C-index to evaluate survival models in the presence of left-truncated data.
@@
 For example, when computing the C-index in the observed sample, the model is evaluated in a test set that suffers from the same bias as the training set.
-Moreover, the similarity in the values of the C-index in the complete sample may be caused by the assumption of no direct correlation between the predictors and latent time to study entry.
+Moreover, the C-indices may be nearly identical in the complete sample because we did not simulate a dependence between the predictors and entry time.
 In other words, the bias in the coefficients is based only on the indirect correlation between the predictors and left truncation induced by (i) the correlation between the predictors and survival and (ii) the correlation between survival and the probability a patient is left-truncated.
@@
 
-Both the small and large $p$ models were designed to highlight left truncation in a real-world dataset and its impact on inference, as opposed to optimizing predictive performance and the prognostic value of predictors.
+Both the small and large $p$ models were designed to highlight left truncation in a real-world dataset and its impact on models, as opposed to optimizing predictive performance and the prognostic value of predictors.
 For this reason, model specifications were relatively simple, e.g.
@@
 In this paper, we implemented an approach for estimating penalized Cox proportional hazards model with LTRC survival data and compared it to penalized models designed for survival data that is right-censored but not left-truncated.
-Using simulation studies and examples from real-world EHR and genomics data, we showed that there is a need for approaches that can both adjust for left truncation and model high dimensional data.
+Using simulation studies and examples from real-world EHR and genomics data, we showed that there is a need for approaches that can both adjust for left truncation and model high-dimensional data.
 In particular, our simulation shows that predictions from models that fail to adjust for left truncation will overestimate true survival probabilities whereas models that properly adjust can yield well calibrated survival predictions, even with high-dimensional data.
-Moreover, in our real-world use case, there were significant differences in coefficient estimates between models that did and did not adjust for left truncation, compromising inference.
+Moreover, in our real-world use case, there were significant differences in coefficient estimates between models that did and did not adjust for left truncation.
 

### 2021-02-08 manuscript.tex
competing interests; apply vs implement

@@
 In some cases survival data is also left-truncated which can give rise to an immortal time bias, but penalized survival methods that adjust for left truncation are not commonly implemented.
-To address these challenges, we implement a penalized Cox proportional hazards model for left-truncated and right-censored survival data and assess implications of left truncation adjustment on bias and interpretation.
+To address these challenges, we apply a penalized Cox proportional hazards model for left-truncated and right-censored survival data and assess implications of left truncation adjustment on bias and interpretation.
 We use simulation studies and a high-dimensional, real-world clinico-genomic database (CGDB) to highlight the pitfalls of failing to account for left truncation in survival modeling.
@@
 
-Here, we implement a penalized Cox proportional hazards model for left-truncated and right-censored (LTRC) survival data and critically assess scenarios in which bias arises from and is exacerbated by left truncation.
+Here, we apply a penalized Cox proportional hazards model for left-truncated and right-censored (LTRC) survival data and critically assess scenarios in which bias arises from and is exacerbated by left truncation.
 We apply the method to simulated data and a real-world use case from a de-identified clinico-genomic database (CGDB).
@@
 \section{Discussion} \label{sec:discussion}
-In this paper, we implemented an approach for estimating penalized Cox proportional hazards model with LTRC survival data and compared it to penalized models designed for survival data that is right-censored but not left-truncated.
+In this paper, we applied an approach for estimating penalized Cox proportional hazards model with LTRC survival data and compared it to penalized models designed for survival data that is right-censored but not left-truncated.
 Using simulation studies and examples from real-world EHR and genomics data, we showed that there is a need for approaches that can both adjust for left truncation and model high-dimensional data.
@@
 We thank T. Hastie and K. Tay for their work in implementing left truncation-adjusted Cox models in \pkg{glmnet}. 
+\textbf{Competing interests}: R.T. and B.N. are paid consultants for Roche.
 

### 2021-04-16 manuscript.tex
unbold text, add last name
de-pluralize Data Sciences (author affils)
Clarifications to the data generating process

@@
 
-%\author[1]{Sarah McGough\footnote{Contributed equally.}}
 \newcommand*\samethanks[1][\value{footnote}]{\footnotemark[#1]}
@@
 \affil[2]{Department of Statistics, Stanford University, Stanford, CA, USA}
-\affil[3]{Department of Biomedical Data Sciences, Stanford University, Stanford, CA, USA}
+\affil[3]{Department of Biomedical Data Science, Stanford University, Stanford, CA, USA}
 
@@
 The observed survival time is given by $Y_i = \min(T_i, U_i)$ and a patient is right-censored if $T_i > U_i$.
-Patients are left-truncated (i.e., hidden from the sample) if $V_i > Y_i$.
+Patients are left-truncated (i.e., excluded from the sample) if $V_i > Y_i$.
 
@@
 The probability, $\pi_k$ that  variable $k$ was equal to $1$ was randomly sampled from a uniform distribution ranging from .2 to .8; that is, we sampled $\pi_k \sim U(0.2, 0.8)$ and set $\mu_k = \Phi^{-1}(\pi_k)$ where $\Phi(\cdot)$ is the cumulative density function of the normal distribution.
-$\Sigma$ was generated by filling a $\tilde{p} \times \tilde{p}$ matrix, $\tilde{\Sigma}$, with $N(0,1)$ random variables and setting  $\Sigma = S/D^TD$ with $S = \tilde{\Sigma}^T\tilde{\Sigma}$ and $D$ a vector containing the diagonal elements of $S$.
+$\Sigma$ was generated by first constructing a positive semi-definite matrix $S = \tilde{\Sigma}^T\tilde{\Sigma}$ by filling the $\tilde{p} \times \tilde{p}$ matrix, $\tilde{\Sigma}$, with $N(0,1)$ random variables.
+$S$ was then scaled into a correlation matrix by setting $\Sigma = S/D^TD$ where $D$ is a vector containing the diagonal elements of $S$.
 
-Latent survival time was simulated from a proportional hazards Weibull distribution with shape parameter $a$ and scale parameter $m_i$ based on evidence from the CGDB as detailed in \autoref{subsec:sim-calibration}.
-Predictors predict latent survival time through their effect on the scale parameter,
+Latent survival time was simulated from a  Weibull distribution based on evidence from the CGDB as detailed in \autoref{subsec:sim-calibration}.
+We used a proportional hazards parameterization of the Weibull distribution with shape parameter $a$, scale parameter $m$, and for $t> 0$, probability density function given by,
+
+\begin{equation}
+f(t) = amt^{a-1}e^{-mt^a}.
+\end{equation}
+
+The predictors were used to model latent survival time through their effect on the scale parameter,
 
@@
 where $\alpha$ is an intercept and $\beta$ is a vector of coefficients.
-$\alpha$ and values of $\beta$ for the CGDB variables were estimated from the CGDB data.
+$\alpha$ and values of $\beta$ for the CGDB variables were estimated from the CGDB data as described in \autoref{subsec:sim-calibration}.
 The probability that the coefficient of a binary variables was non-zero was $0.5$ and drawn from a Bernoulli distribution.
@@
 The full coefficient vector combined the coefficients from the CGDB variables with the coefficients from the simulated binary variables.
-The scale parameter was set to $m = \alpha + x^T\beta - E(x^T\beta)$ so that predicted survival averaged across patients was equal to predicted survival from the intercept only Weibull model.
+The natural logarithm of the scale parameter was set to $\log(m_i) = \alpha + x^T_i\beta - E(x^T_i\beta)$ so that predicted survival averaged across patients was equal to predicted survival from the intercept only Weibull model.
 
@@
 We thank T. Hastie and K. Tay for their work in implementing left truncation-adjusted Cox models in \pkg{glmnet}. 
-\textbf{Competing interests}: R.T. and B.N. are paid consultants for Roche.
+Competing interests: R. Tibshirani and B. Narasimhan are paid consultants for Roche.
 

### 2021-04-16 R/calibrate_sim.R
Modifications to simulation code so that it can be run without required the underlying data

@@ -9,6 +9,7 @@ fit_survregs <- function(f, data){
   }
   fits$survspline <- flexsurv::flexsurvspline(f, data = data, k = 1)
   fits$survspline$pretty_dist <- "Spline"
+  
   return(fits)
 }
 
@@ -93,7 +94,6 @@ make_survregs <- function(survregfits, coxfit){
     fits[[i]]$coefs <- add_coef(survregfits[[i]], stats::coef(coxfit))
     fits[[i]]$knots <- survregfits[[i]]$knots  
     fits[[i]]$dlist <- survregfits[[i]]$dlist
-    browser()
   }
   names(fits) <- names(survregfits)
   return(fits)
@@ -108,8 +108,27 @@ plot_cumhaz_rc_os <- function(cumhaz_rc, cumhaz_os) {
   plot_cumhaz(pdata, grp = "facet")
 }
 
+remove_model_data <- function(object, ...) {
+  UseMethod("remove_model_data")
+}
+
+remove_model_data.default <- function(object, ...) {
+  return(object)
+}
+
+remove_model_data.flexsurvreg <- function(object, ...) {
+  object$data$Y <- NULL
+  object$data$m <- NULL
+  # Model has no covariates so we can keep object$data$ml
+  return(object)
+}
+
+remove_model_data.list <- function(object, ...) {
+  lapply(object, remove_model_data)
+}
+
 #' @export
-calibrate_sim <- function(f, data){
+calibrate_sim <- function(f, data, store_x = TRUE){
   xy <- make_xy(data, f)
   
   make_f <- function(f_lhs, f_rhs){
@@ -124,7 +143,7 @@ calibrate_sim <- function(f, data){
     ltrc = survival::survfit(make_f(f_os2, ~1), data)
   )
   os_base <- fit_survregs(make_f(f_os2, ~1), data)
-  os_cox <- survival::coxph(make_f(f_os2, f), data)
+  os_cox <- survival::coxph(make_f(f_os2, f), data, x = FALSE, y = FALSE)
   os_comparisons <- compare_survregs(os_base, os_km$ltrc)
   
   # Right censoring
@@ -135,11 +154,12 @@ calibrate_sim <- function(f, data){
     ltrc = survival::survfit(make_f(f_rc2, ~1), data) 
   )
   rc_base <- fit_survregs(make_f(f_rc2, ~1), data)
-  rc_cox <- survival::coxph(make_f(f_rc2, f), data) 
[…]

### 2021-04-16 R/run_sim.R

@@ -21,9 +21,15 @@ sim_vars <- function(n_pats, K){
 #' @export
 sim_x <- function(n_pats = 1000, sim_settings, p_bin) {
   # RWD
-  rows_to_sample <- sample.int(n = nrow(sim_settings$xy$x), 
-                               size = n_pats, replace = TRUE)
-  X <- sim_settings$xy$x[rows_to_sample, ]
+  if (!is.null(sim_settings[["x"]])) { # Bootstrap observed data if exists
+    rows_to_sample <- sample.int(n = nrow(sim_settings$x), 
+                                 size = n_pats, replace = TRUE)
+    X <- sim_settings$x[rows_to_sample, ]
+  } else { # Otherwise sample from multivariate normal distribution
+    X <- mvtnorm::rmvnorm(n_pats, mean = sim_settings$x_mean, 
+                          sigma = sim_settings$x_vcov)
+  }
+
   n_rwdvars <- ncol(X)
   rwd_type <- rep("RWD", ncol(X))
   attr(X, "type") <- rwd_type

### 2021-04-16 manuscript.tex

--- 2021-02-08 manuscript.tex
@@
 
-Simon et. al\cite{simon2011regularization} introduced a fast, pathwise algorithm to fit a regularized Cox proportional hazards model solving for
+Simon \emph{et al.} \cite{simon2011regularization} introduced a fast, pathwise algorithm to fit a regularized Cox proportional hazards model solving for
 
@@
 
-where $\beta$ is the vector of regression coefficients and is maximized over \autoref{eqn:simon} subject to a constraint defined by the penalty $\lambda \cdot P_{\alpha}\beta$.
-As described elsewhere \cite{simon2011regularization}, the penalty can take the form of the $l_1$ (lasso), $l_2$ (ridge), or elastic net penalty,  and the algorithm fits over a path of $\lambda$ values via cyclical coordinate descent.
+where $\beta$ is the vector of regression coefficients and is maximized over \autoref{eqn:simon} subject to a constraint defined by $\lambda \cdot P_{\alpha}\beta$, and $P_{\alpha}\beta$ is the $l_1$ (lasso), $l_2$ (ridge), or elastic net penalty on the sum of coefficients $\beta$ described elsewhere \cite{simon2011regularization, tibshirani1996regression, tikhonov1963ridge, zou2005regularization}.
+The algorithm by Simon \emph{et al.} fits over a path of $\lambda$ values via cyclical coordinate descent.
 Instead of solving a least squares problem at each iteration, the algorithm solves a penalized reweighted least squares problem, in which observation weights $w$ are defined in the coordinate descent least squares minimization step.
@@
 In some cases the adjustment can have a considerable impact on the hazard ratios, often moving a large hazard ratio all the way to $1$ (i.e., to a coefficient of $0$).
-This is most notable with diagnosis year, which has a hazard ratio of approximately $1.5$ in both the small and large $p$ models but is set to $1$ after the left truncation adjustment.
+This is most notable with diagnosis year, which has a hazard ratio of approximately $1.5$ in both the small and large $p$ models but is reduced to $1$ or less than $1$ after the left truncation adjustment.
 This effect is likely due to the strong negative correlation between calendar time and the probability that a patient is left-truncated.
@@
 \autoref{tbl:model-discrimination} compares models using the C-index.
-The C-index is relatively low in all of these illustrative models, though is artificially higher in the unadjusted models that do not account for left truncation---sometimes considerably so---and decreases after adjustment as the predictor hazard ratios are pushed toward $1$ and there are fewer differences in predicted relative risks across patients.
+The C-index is relatively low in all of these illustrative models, though is artificially higher in the unadjusted models that do not account for left truncation---sometimes considerably so---and decreases after adjustment as there are fewer differences in predicted relative risks across patients and the discriminatory ability of the model is low.
 As expected, the lasso model has a higher C-index in the larger $p$ setting as the Cox model is considerably overfitted.
@@
 In our real-world use case, an analyst could have mistakenly concluded that the high C-index in the non-adjusted model was indicative of good model performance when it was in fact driven primarily by a high correlation between predictors and left truncation.
-For instance, in our large $p$ model the hazard ratio for the predictor "year of diagnosis" moved from being the largest in absolute value to $1$ after adjusting for left truncation.
-The high C-index in the non-adjusted model was artificial because patients diagnosed earlier appeared to survive longer due to immortal time bias.
-To facilitate model interpretation, we recommend that models should be evaluated using both discrimination and calibration based metrics.
+For instance, in our large $p$ model the hazard ratio for the predictor "year of diagnosis" moved from being the largest in absolute value to around $1$ after adjusting for left truncation.
+The strongly positive effect of "year of diagnosis" on the hazard in the non-adjusted model was artificial because patients diagnosed earlier appeared to survive longer due to immortal time bias.
+This effect is reversed completely after the left truncation adjustment, and illustrates the bias to coefficient estimates that is driven by the association between predictors and left truncation time.
+In the case of "year of diagnosis," this association is negative and drives an upward bias to the hazard ratio.
+While many of the other hazard ratios in the large $p$ model were pushed towards $1$ after left truncation adjustment, the direction of this bias depends on the direction of the association with left truncation time and varies by predictor.
+Because the high C-index misleadingly suggested good model performance, we recommend that models should be evaluated using both discrimination and calibration based metrics to facilitate interpretation.
 

### 2021-04-16 main.R
Update main.R so it can run using either prestored simulation settings or can calibrate the simulation setting itself

@@ -18,7 +18,7 @@ library("rngtools") # Required by dorng
 library("rsample")
 library("survival")
 library("tidyr") # (>= 1.0.0) for pivot_longer() and pivot_wider()
-library("xfun") # (>= 0.13.0) for cache_rds()
+# library("xfun") # We will use xfun::cache_rds() below, but not attach the package here
 library("xtable")
 
 # Sourced R files
@@ -112,7 +112,11 @@ rm(train_data)
 rm(test_data)
 
 # Calibrate simulation ---------------------------------------------------------
-sim_settings <- calibrate_sim(f_small, data = data)
+if (file.exists("sim_settings.rds")) { # Interested readers can run this
+  sim_settings <- readRDS("sim_settings.rds")
+} else { # This setting is run for the paper
+  sim_settings <- calibrate_sim(f_small, data = data, store_x = TRUE)
+}
 
 # Save cumulative hazard plot
 ggsave("figs/sim_calibration_cumhaz.pdf", 

### 2021-04-19 manuscript.tex
minor edits- Efron approx, HR bias

--- 2021-04-16 manuscript.tex
@@
 Tied survival times are handled via the Breslow approximation \cite{breslow1972} and details provided in \cite{simon2011regularization}.
+Note that the computationally efficient Efron approximation will be added to the next version of \pkg{glmnet}.
 
@@
 The strongly positive effect of "year of diagnosis" on the hazard in the non-adjusted model was artificial because patients diagnosed earlier appeared to survive longer due to immortal time bias.
-This effect is reversed completely after the left truncation adjustment, and illustrates the bias to coefficient estimates that is driven by the association between predictors and left truncation time.
-In the case of "year of diagnosis," this association is negative and drives an upward bias to the hazard ratio.
+This effect is reversed completely after the left truncation adjustment, and illustrates that biases to coefficient estimates are driven by the association between predictors and left truncation time.
+In the case of "year of diagnosis," this association is negative and biases the hazard ratio upwards.
 While many of the other hazard ratios in the large $p$ model were pushed towards $1$ after left truncation adjustment, the direction of this bias depends on the direction of the association with left truncation time and varies by predictor.
-Because the high C-index misleadingly suggested good model performance, we recommend that models should be evaluated using both discrimination and calibration based metrics to facilitate interpretation.
+In cases where predictors and left truncation time are positively associated, hazard ratios are biased downwards because the predictors are associated with more immortal time.
 

### 2021-04-22 R/calibrate_sim.R
Only keep object elements truly needed for simulations

@@ -108,25 +108,6 @@ plot_cumhaz_rc_os <- function(cumhaz_rc, cumhaz_os) {
   plot_cumhaz(pdata, grp = "facet")
 }
 
-remove_model_data <- function(object, ...) {
-  UseMethod("remove_model_data")
-}
-
-remove_model_data.default <- function(object, ...) {
-  return(object)
-}
-
-remove_model_data.flexsurvreg <- function(object, ...) {
-  object$data$Y <- NULL
-  object$data$m <- NULL
-  # Model has no covariates so we can keep object$data$ml
-  return(object)
-}
-
-remove_model_data.list <- function(object, ...) {
-  lapply(object, remove_model_data)
-}
-
 #' @export
 calibrate_sim <- function(f, data, store_x = TRUE){
   xy <- make_xy(data, f)
@@ -154,26 +135,33 @@ calibrate_sim <- function(f, data, store_x = TRUE){
     ltrc = survival::survfit(make_f(f_rc2, ~1), data) 
   )
   rc_base <- fit_survregs(make_f(f_rc2, ~1), data)
-  rc_cox <- survival::coxph(make_f(f_rc2, f), data, x = FALSE, y = FALSE) 
   rc_comparisons <- compare_survregs(rc_base, rc_km$ltrc)
   
   # Combined overall survival and right censoring
   cumhaz_plot <- plot_cumhaz_rc_os(rc_comparisons$cumhaz, os_comparisons$cumhaz)
-
   
-  # Return
+  # Create object to return
+  ## Only keep information from "flexsurvreg" objects that is needed 
+  ## (since we can't share raw data)
+  subset_flexsurvreg <- function (x) {
+    lapply(x, function (z) {
+      return(z[c("dlist","res.t")]) 
+    })
+  }
+  os_base <- subset_flexsurvreg(os_base)
+  rc_base <- subset_flexsurvreg(rc_base)
+  
+  
+  ## Create the list
   res <- list(os_km = os_km,
               os_base = os_base,
-              os_cox = os_cox,
+              os_coef = stats::coef(os_cox),
               os_comparisons = os_comparisons,
               rc_base = rc_base,
-              rc_cox = rc_cox,
               rc_comparisons = rc_comparisons,
               cumhaz_plot = cumhaz_plot,
               formula = f)
   
-  ## Remove data from any models because we can't share it
-  res <- lapply(res, remove_model_data)
   
[…]

### 2021-04-22 R/run_sim.R

@@ -47,7 +47,6 @@ sim_x <- function(n_pats = 1000, sim_settings, p_bin) {
 set_beta <- function(X, sim_settings, dist) {
   
   basefit <- sim_settings$os_base[[dist]]
-  coxfit = sim_settings$os_cox
   
   type <- attr(X, "type")
   beta <- rep(NA, length(type))
@@ -57,7 +56,7 @@ set_beta <- function(X, sim_settings, dist) {
   rwd_index <- which(type == "RWD")
   name <- basefit$dlist$location
   pos <- which(basefit$dlist$pars == name)
-  beta[rwd_index] <- stats::coef(coxfit)
+  beta[rwd_index] <- sim_settings$os_coef
   
   ## Binary variables
   which_bin <- which(type == "Binary")
@@ -98,7 +97,7 @@ predict_location <- function(intercept, beta = NULL, X = NULL, inv_transforms){
 }
 
 surv_rng <- function(n, beta, basefit, X){
-  
+
   # Location parameter
   loc_pos <- which(basefit$dlist$pars == basefit$dlist$location)
   loc_name <- basefit$dlist$pars[loc_pos]

### 2021-04-23 02-simulation.R
Some restructuring so that the simulations can be run without access to the raw data

@@ -1,0 +1,89 @@
+# Calibrate simulation ---------------------------------------------------------
+if (file.exists("sim_settings.rds")) { # Interested readers can run this
+  sim_settings <- readRDS("sim_settings.rds")
+} else { # This setting is run for the paper
+  data <- readRDS("data.rds")
+  sim_settings <- calibrate_sim(f_small, data = data, store_x = TRUE)
+}
+
+# Save cumulative hazard plot
+ggsave("figs/sim_calibration_cumhaz.pdf", 
+       sim_settings$cumhaz_plot,
+       height = 5, width = 10)
+
+# Example simulated data -------------------------------------------------------
+# No predictors (intercept only model)
+params <- set_params(sim_settings = sim_settings, dist = "weibullPH")
+simdata <- sim_survdata(params = params, n_pats = 10000)
+simdata_summary_int <- summarize_simdata(simdata, km_rwd = sim_settings$os_km$rc,  
+                                         save = TRUE, name = "int")
+
+# Now include predictors (and hence create informative censoring when
+# using the Kaplan-Meier estimator). The design matrix X and parameters will 
+# be used in the simulations that follow
+x_sim <- sim_x(n_pats = 5000, sim_settings, p_bin = 0)
+params <- set_params(x_sim, sim_settings, dist = "weibullPH")
+simdata <- sim_survdata(x_sim, params)
+simdata_summary_p11 <- summarize_simdata(simdata, km_rwd = sim_settings$os_km$rc,
+                                         save = TRUE, name = "p11")
+
+# Run simulation for unpenalized Cox model -------------------------------------
+sim_coxph_p21 <- xfun::cache_rds({
+  run_sim(n_sims = N_SIMS, x = x_sim, params = params, method = "coxph")
+}, file = "sim_coxph_p21.rds", rerun = RERUN_CACHE)
+sim_coxph_21_summary <- summarize_sim(sim_coxph_p21, save = TRUE,
+                                      model_name = "coxph_p21")
+rm(sim_coxph_p21)
+
+# Run simulation for lasso model with lambda = 0 and small p -------------------
+# run_sim1(simdata, method = "coxnet", lambda = 0) # For debugging
+sim_coxlasso_lambda0_p21 <- xfun::cache_rds({
[…]

### 2021-04-23 R/calibrate_sim.R

@@ -145,7 +145,7 @@ calibrate_sim <- function(f, data, store_x = TRUE){
   ## (since we can't share raw data)
   subset_flexsurvreg <- function (x) {
     lapply(x, function (z) {
-      return(z[c("dlist","res.t")]) 
+      return(z[c("dlist", "res.t")]) 
     })
   }
   os_base <- subset_flexsurvreg(os_base)

### 2021-04-23 main.R

@@ -1,433 +1,4 @@
-# Setup ------------------------------------------------------------------------
-rm(list = ls())
-
-# Packages
-library("data.table")
-library("doParallel")
-library("doRNG") # Reproducible parallel foreach loops
-library("dplyr") 
-library("foreach")
-library("glmnet")
-library("ggplot2")
-library("ggpubr")
-library("impute") # From bioconductor
-library("knitr")
-library("parallel")
-library("purrr")
-library("rngtools") # Required by dorng
-library("rsample")
-library("survival")
-library("tidyr") # (>= 1.0.0) for pivot_longer() and pivot_wider()
-# library("xfun") # We will use xfun::cache_rds() below, but not attach the package here
-library("xtable")
-
-# Sourced R files
-source("R/plot_patient_followup.R")
-source("R/impute_genes.R")
-source("R/make_xy.R")
-source("R/calibrate_sim.R")
-source("R/run_sim.R")
-source("R/adjust_surv.R")
-source("R/tidycoef.R")
-source("R/concordance.R")
-source("R/calibrate.R")
-
-# Settings
-set.seed(77)
-theme_set(theme_bw())
-center_title <- function() theme(plot.title = element_text(hjust = 0.5))
-N_SIMS <- 200
-
-# Caching
-RERUN_CACHE <- TRUE # Set to TRUE to rerun all cached results
-if (!dir.exists("cache")){
-  dir.create("cache")
-}
-
-# Parallel
-PARALLEL <- TRUE
-if (PARALLEL) {
-  cl <- parallel::makeCluster(20, setup_strategy = "sequential", outfile = "simout")
-  registerDoParallel(cl)
-}
-
-# Store numbers to use in text of model documentation
-txt <- list() # List for in-line text statistics
-
-# Followup plot ----------------------------------------------------------------
-p_followup <- plot_patient_followup()
-ggsave("figs/followup.pdf", p_followup, height = 5, width = 7)
-
-# Load data --------------------------------------------------------------------
-data <- readRDS("data.rds")
-
-# Split data -------------------------------------------------------------------
-# Using rsample package
-data_split <- initial_split(data, prop = .75)
-train_data <- training(data_split)
-test_data <- testing(data_split)
-
-# Clean up
-rm(data_split)
-
[…]

### 2021-04-23 main.R
Fix to main.R

@@ -1,4 +1,4 @@
 source("01-setup.R")
-#source("02-simulation.R")
+source("02-simulation.R")
 source("03-application.R")
 source("04-txtstats.R")

### 2021-04-27 manuscript.tex
acknowledge reviewers
A few text updates + recompile manuscript with updated results

--- 2021-04-16 manuscript.tex
@@
 The adjusted models tend to be better calibrated for patients with the lowest survival probabilities but poorly calibrated for patients with higher survival probabilities.
-One possible reason for this result is that the adjusted models pushed many of the hazard ratios to $1$ and do not adequately separate low and high risk patients.
+One possible reason for this result is that the adjusted models pushed many of the hazard ratios to $1$ and did not adequately separate low and high risk patients.
 Probability rescaling such as Platt scaling or isotonic regression may be effective on the adjusted probabilities here, which resemble the characteristic sigmoid-shaped calibration curves produced by different learning algorithms, including the lasso.
@@
 In our real-world use case, an analyst could have mistakenly concluded that the high C-index in the non-adjusted model was indicative of good model performance when it was in fact driven primarily by a high correlation between predictors and left truncation.
-For instance, in our large $p$ model the hazard ratio for the predictor "year of diagnosis" moved from being the largest in absolute value to around $1$ after adjusting for left truncation.
-The strongly positive effect of "year of diagnosis" on the hazard in the non-adjusted model was artificial because patients diagnosed earlier appeared to survive longer due to immortal time bias.
-This effect is reversed completely after the left truncation adjustment, and illustrates the bias to coefficient estimates that is driven by the association between predictors and left truncation time.
-In the case of "year of diagnosis," this association is negative and drives an upward bias to the hazard ratio.
-While many of the other hazard ratios in the large $p$ model were pushed towards $1$ after left truncation adjustment, the direction of this bias depends on the direction of the association with left truncation time and varies by predictor.
+For instance, in our large $p$ model the hazard ratio for the predictor ``year of diagnosis'' moved from being the largest in absolute value to around 1 after adjusting for left truncation.
+The strong positive relationship between ``year of diagnosis'' on the hazard in the non-adjusted model was artificial because patients diagnosed earlier appeared to survive longer due to immortal time bias.
+This relationship reversed completely after adjusting for left truncation, and illustrates that biases to coefficient estimates are driven by the association between predictors and left truncation time.
+In the case of ``year of diagnosis'', this association was negative and biased the hazard ratio upwards.
+While many of the other hazard ratios in the large $p$ model were pushed towards 1 after left truncation adjustment, the direction of this bias depends on the direction of the association with left truncation time and varies by predictor.
+In cases where predictors and left truncation time are positively associated, hazard ratios are biased downwards because the predictors are associated with more immortal time.
 Because the high C-index misleadingly suggested good model performance, we recommend that models should be evaluated using both discrimination and calibration based metrics to facilitate interpretation.

### 2021-04-27 manuscript.tex

@@
 \section*{Acknowledgments} 
-We thank T. Hastie and K. Tay for their work in implementing left truncation-adjusted Cox models in \pkg{glmnet}. 
+We thank colleagues at Flatiron Health, Inc.
+and Foundation Medicine, Inc.
+and the anonymous reviewers for their thoughtful reviews and constructive feedback to improve the quality of this work.
+We thank T.
+Hastie and K.
+Tay for their work in implementing left truncation-adjusted Cox models in \pkg{glmnet}.
 Competing interests: R. Tibshirani and B. Narasimhan are paid consultants for Roche.

### 2021-04-29 manuscript.tex
Minor wording edit
tweak Efron statement

@@
 Instead of solving a least squares problem at each iteration, the algorithm solves a penalized reweighted least squares problem, in which observation weights $w$ are defined in the coordinate descent least squares minimization step.
-The optimal $\lambda$ can be identified through cross-validation.
+The parameter $\lambda$ can be identified through cross-validation.
 Tied survival times are handled via the Breslow approximation \cite{breslow1972} and details provided in \cite{simon2011regularization}.
-Note that the computationally efficient Efron approximation will be added to the next version of \pkg{glmnet}.
+Note that the Efron approximation will be added to the next version of \pkg{glmnet}.
 

### 2021-04-29 manuscript.tex
data are!

@@
 Statistical models that attempt to study the effects of many predictors on survival typically implement feature selection or penalized methods to mitigate the undesirable consequences of overfitting.
-In some cases survival data is also left-truncated which can give rise to an immortal time bias, but penalized survival methods that adjust for left truncation are not commonly implemented.
+In some cases survival data are also left-truncated which can give rise to an immortal time bias, but penalized survival methods that adjust for left truncation are not commonly implemented.
 To address these challenges, we apply a penalized Cox proportional hazards model for left-truncated and right-censored survival data and assess implications of left truncation adjustment on bias and interpretation.

### 2021-04-30 manuscript.tex
clarify LT time / entry time

@@
 However, patients who die before reaching the milestone are left-truncated (in red), and only those who have survived until eligibility (in black) are observed.
-Left truncation time, or the time between diagnosis and cohort entry, is shown with a dashed line.}
+Left truncation time, or the time between diagnosis and cohort entry, is shown with a dashed line.
+Left truncation time is also referred to as ``entry time.''}
 \label{fig:followup}
@@
 Let individual $i$ have survival data of the form $(v_i, y_i, e_i, x_i)$, where individual $i$ enters the study at time $v_i$.
+Time $v_i$ is the left truncation time, also referenced in this text as entry time.
 The left-truncated risk set $R_{i\ast} = \{ j : y_j \geq t_i > v_j \}$.
 In other words, subjects are only considered at risk at a given event time after they have been observed to be at risk (i.e.
-exceeded their left truncation time, shown with a solid black line in \autoref{fig:followup} ).
+exceeded their left truncation time / passed milestone entry time, shown with a solid black line in \autoref{fig:followup} ).
 For example, at the time of death of Patient 3 in \autoref{fig:followup} , only Patient 4 has reached the milestone for study entry and is at risk.
@@
 \includegraphics[max size={\textwidth}]{figs/left_trunc_hist.pdf} 
-\caption{Distribution of left truncation time (days) in non-small cell lung cancer patients in the CGDB}
+\caption{Distribution of left truncation time (days) in non-small cell lung cancer patients in the CGDB.}
  \label{fig:left-trunc-time}

### 2021-04-30 manuscript.tex
footnote change
minor rewordings

@@
 Patients are followed until death (open circle) or censoring (cross).
-However, patients who die before reaching the milestone are left-truncated (in red), and only those who have survived until eligibility (in black) are observed.
+However, patients who die or are censored before reaching the milestone are left-truncated (in red), and only those who have survived until eligibility (in black) are observed.
 Left truncation time, or the time between diagnosis and cohort entry, is shown with a dashed line.
@@
 The CGDB is a US-based de-identified oncology database that combines real-world, patient-level clinical data and outcomes with patient-level genomic data to provide a comprehensive patient and tumor profile.
-The de-identified data originated from approximately 280 US cancer clinics (~800 sites of care).
+The de-identified data originated from approximately 280 US cancer clinics (roughly 800 sites of care).
 Patients in the CGDB have received at least one Foundation Medicine comprehensive genomic profiling test as well as have had their electronic health records captured by Flatiron Health.
@@
 \end{tabularx}
-\scriptsize Note: ``Adjusted'' models incorporate left truncation times.
+\scriptsize Note: ``Adjustment'' indicates models with a left truncation adjustment.
 \end{threeparttable}
@@
 One possible reason for this result is that the adjusted models pushed many of the hazard ratios to $1$ and did not adequately separate low and high risk patients.
-Probability rescaling such as Platt scaling or isotonic regression may be effective on the adjusted probabilities here, which resemble the characteristic sigmoid-shaped calibration curves produced by different learning algorithms, including the lasso.
+Probability rescaling such as Platt scaling or isotonic regression may be effective on the adjusted probabilities here, which resemble the characteristic sigmoid-shaped calibration curves produced by different learning algorithms, including the lasso; however the implementation is less straightforward in a survival context \cite{goldstein2021xcal}.
 
@@
 
-In our real-world use case, an analyst could have mistakenly concluded that the high C-index in the non-adjusted model was indicative of good model performance when it was in fact driven primarily by a high correlation between predictors and left truncation.
+In our real-world use case, an analyst could have mistakenly concluded that the C-index in the non-adjusted model was indicative of good model performance when it was in fact driven in part by a high correlation between predictors and left truncation.
 For instance, in our large $p$ model the hazard ratio for the predictor ``year of diagnosis'' moved from being the largest in absolute value to around 1 after adjusting for left truncation.

### 2021-04-30 manuscript.tex
minor edit

@@
 
-The extension of penalized regression methods to survival data have been extensively described and applied, for example in microarray gene expression data \cite{gui2005penalized}, transcriptomes \cite{wu2011penalized}, and injury scale codes \cite{mittal2013penalized}.
+The extensions of penalized regression methods to survival data have been extensively described and applied, for example in microarray gene expression data \cite{gui2005penalized}, transcriptomes \cite{wu2011penalized}, and injury scale codes \cite{mittal2013penalized}.
 Like linear regression, the performance and fit of a Cox proportional hazards model is known to deteriorate with large $p$, providing an advantage to penalized approaches in high-dimensional, time-to-event feature sets.

### 2021-06-17 example.Rmd
having trouble knitting!

---
title: "Example penalized LTRC survival model with simulated data"
author: "McGough et al"
date: "6/17/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This is a brief guide to implementing a penalized survival model for left-truncated and right-censored data (LTRC), described in McGough *et al.* 2021. This vignette uses simulated data based on the Foundation Medicine and Flatiron Health Clinico-Genomic Database (CGDB), as described in the paper [repository](https://github.com/phcanalytics/coxnet-ltrc).

## Load Data and Setup

The simulated data of interest, `simdata`, contains 5000 patients and 18 variables.

```{r}
# Setup and Simulations --------------------------------------------------------

library(dplyr) 
library(glmnet)
library(ggplot2)
library(knitr)
library(rsample)
library(rmarkdown)

source("R/run_sim.R")
source("R/calibrate.R")

# Read data
sim_settings <- readRDS("sim_settings.rds")
params <- set_params(sim_settings = sim_settings, dist = "weibullPH")
simdata <- sim_survdata(params = params, n_pats = 10000)

head(simdata)
```

## Prep Model Data

Here we randomly split the simulated data, `simdata`, into train (75%) and test (25%) data. 
We also only use patients from the simulated data who are alive at entry time (`death_time` >= `entry_time`, approximately 44% of patients).
Columns 7:17 contain our predictors of interest.

```{r}
# Filter to patients alive at study entry --------------------------------------
simdata_sub <- simdata %>% filter(hidden==0)

# Split data -------------------------------------------------------------------
# Using rsample package
set.seed(7)
data_split <- initial_split(simdata_sub, prop = .75)
train_data <- training(data_split)
test_data <- testing(data_split)

# Clean up
rm(data_split)
```

## Run Model

Here we run a simple example penalized Lasso model for LTRC using `cv.glmnet()` in `glmnet` > 4.1. The lambda which minimizes the partial likelihood deviance (`lambda.min`) is chosen through 5-fold cross-validation. For more details, see the [glmnet homepage](https://glmnet.stanford.edu/articles/Coxnet.html).

```{r}
# Prep x and y matrices - train
x_train <- train_data %>% select(7:17) %>% as.matrix()
y_train <- Surv(train_data$entry_time,
          train_data$death_time,
          train_data$dead)

# Repeat for test data
x_test <- test_data %>% select(7:17) %>% as.matrix()
y_test <- Surv(test_data$entry_time,
          test_data$death_time,
          test_data$dead)

# 5-fold CV glmnet
mod <- cv.glmnet(x_train, y_train, family="cox", nfolds = 5)

```

## Model Coefficients

Choosing the best regularized model using `lambda.min`, we can assess model coefficients, which represent the log hazard ratios:
```{r}
coef(mod, s = "lambda.min")
```

Or alternatively, we may choose the model which minimizes deviance within 1 standard deviation of the minimum `lambda.1se`, which results in a sparser model:

```{r}
coef(mod, s = "lambda.1se")
```

The full path of regularization can also be plotted:
```{r}
plot(mod)
```

## Model Predictions and Performance

We will make out-of-sample predictions on the 25% hold-out set using `lambda.min`. 
```{r}
predictions <- predict(mod, x_test, s="lambda.min")
head(predictions)
```

### C-index

```{r}
get_cindex <- function(newx, newy, fit, lambda){
  survival::concordance(newy ~ predict(fit, newx = newx, s = lambda), reverse=TRUE)
}

get_cindex(x_test, y_test, mod, "lambda.min")
```

### Calibration

```{r}
# Predict survival probabilities
survfit <- survfit(mod, newx = x_test, s="lambda.min", x= x_train, y= y_train)

# Get calibration at time points
cal <- calibrate.survfit(object = survfit, times = seq(.5, 3, .5), y = y_test)

autoplot.calibrate(cal)
```

### 2021-06-18 01-setup.R
Needed to update 01-setup.R with last commit

@@ -23,7 +23,7 @@ source("R/plot_patient_followup.R")
 source("R/impute_genes.R")
 source("R/make_xy.R")
 source("R/calibrate_sim.R")
-source("R/run_sim.R")
+source("R/sim.R")
 source("R/adjust_surv.R")
 source("R/tidycoef.R")
 source("R/concordance.R")

### 2021-06-18 example.Rmd
Updates to incorporate example analysis

@@
 title: "Example penalized LTRC survival model with simulated data"
-author: "McGough et al"
-date: "6/17/2021"
-output: pdf_document
+author: "Sarah F. McGough, Devin Incerti, Svetlana Lyalina, Ryan Copping, Balasubramanian Narasimhan, Robert Tibshirani"
+date: "`r Sys.Date()`"
+output: 
+  html_document:
+    toc: yes
+    toc_depth: 2
+    toc_float: true
+    number_sections: TRUE
 ---
@@
 ```{r setup, include=FALSE}
+# So that it can be viewed on GitHub pages, render with:
+# rmarkdown::render(input = "example.Rmd", output_dir = "docs")
 knitr::opts_chunk$set(echo = TRUE)
@@
 
-## Introduction
+# Introduction
+This guide shows how to use penalized survival models for prediction with left-truncated and right-censored data (LTRC) as described in McGough *et al.* 2021.
+The analyses utilize a simulated dataset based on the Foundation Medicine and Flatiron Health Clinico-Genomic Database (CGDB), as described in the paper.
 
-This is a brief guide to implementing a penalized survival model for left-truncated and right-censored data (LTRC), described in McGough *et al.* 2021.
-This vignette uses simulated data based on the Foundation Medicine and Flatiron Health Clinico-Genomic Database (CGDB), as described in the paper [repository](https://github.com/phcanalytics/coxnet-ltrc).
+We will use `glmnet` to fit the models, `rsample` to split the data into training and test sets, and various features of the `survival` package.
+We will also need a script containing custom functions for calibrating survival predictions.
 
-## Load Data and Setup
+```{r, message = FALSE, warning = FALSE}
+# Packages used in this script
+library("glmnet")
+library("rsample")
+library("survival")
 
-The simulated data of interest, `simdata`, contains 5000 patients and 18 variables.
+# Custom functions for calibration
+library("data.table") # Dependency for calibration functions
+library("ggplot2") # Required to visualize calibrations
+source("R/calibrate.R") # Our custom calibration functions
+theme_set(theme_bw()) # ggplot theme
 
-```{r}
-# Setup and Simulations --------------------------------------------------------
+# There is some randomness in the analysis include splitting
+# the data into train/test sets and cross validation.
+set.seed(7) 
+```
 
-library(dplyr) 
-library(glmnet)
-library(ggplot2)
-library(knitr)
-library(rsample)
-library(rmarkdown)
+# LTRC data
+While we are not permitted to share data from the CGDB, we can simulate data that is consistent with what we observe.
+Our simulated dataset is contained in the file `simdata.rds`.
+Details on how we simulated the data are described in the paper.
+If you would like to see the code used to simulate the data you can view/run the file `simdata.R`.
 
-source("R/run_sim.R")
-source("R/calibrate.R")
-
-# Read data
-sim_settings <- readRDS("sim_settings.rds")
-params <- set_params(sim_settings = sim_settings, dist = "weibullPH")
-simdata <- sim_survdata(params = params, n_pats = 10000)
-
+```{r simdata}
+simdata <- as.data.frame(readRDS("simdata.rds"))
 head(simdata)
@@
 
-## Prep Model Data
+There are two characteristics of the data that are artificial and would not be observed in reality.
+First, we simulate (latent) `death_time`, but deaths are only observed if they occur prior to censoring; that is, since the data is right-censored, we only observe `event_time`.
+Second, some patients are not included in the data (i.e., they are `hidden`), because their `event_time` occurs prior to the time they would have entered the dataset (their `entry_time`).
+These patients are left-truncated.
 
-Here we randomly split the simulated data, `simdata`, into train (75%) and test (25%) data. 
-We also only use patients from the simulated data who are alive at entry time (`death_time` >= `entry_time`, approximately 44% of patients).
-Columns 7:17 contain our predictors of interest.
+To create a realistic LTRC dataset we will consequently remove the latent `death_time` variable and subset the data to patients that were not left-truncated.
+
+```{r simdataAfterLT}
+simdata <- simdata[simdata$hidden == 0, ]
+simdata$death_time <- NULL
+```
+
+## Train and test sets
+For prediction problems, it is common to split the data into training and tests sets.
+Here we use the `rsample` package to randomly split the data so that 75% of patients are in the training set and 25% are in the test set.
+
+```{r split}
+data_split <- initial_split(simdata, prop = .75)
+train_data <- training(data_split)
+test_data <- testing(data_split)
+```
+
+# Model fitting
+## Model data
+### Input matrix
+`glmnet` requires an input matrix `x`, which which must be a `matrix` object. Here we consider a simple example with a small number of predictors contained in `x`.
+
+```{r xvars}
+x_vars <- attr(simdata, "x_vars")
+x_vars
+```
+
+We create `x` matrices for both the training and the test sets.
+
+```{r xData}
+x_train <- as.matrix(train_data[, x_vars])
+x_test <- as.matrix(test_data[, x_vars])
+```
+
+### Response
+When fitting Cox proportional hazards models with `glmnet`, the response should be a `survival::Surv` object.
+The key difference between modeling LTRC and RC data, is that the response must contain start-stop intervals when using LTRC data.
+Start-stop `Surv` object are created by passing the time that a patient enters the dataset (`time`), the follow-up time (`time2`), and a status indicator equal to 1 if a patient has died and 0 if they are alive.
+
+As with `x`, we create `y` for both the training and the test sets.
+
+```{r trainTestData}
+y_train <- Surv(time = train_data$entry_time,
+                time2 = train_data$event_time,
+                event = train_data$dead)
+
+y_test <- Surv(test_data$entry_time,
+               test_data$event_time,
+               test_data$dead)
+```
+
+## Cross validation
+We will fit a penalized Lasso model appropriate for LTRC data using `cv.glmnet()` (note: requires `glmnet` > 4.1).
+The value of the tuning parameter$\lambda$ that minimizes the partial likelihood deviance (`lambda.min`) is selected using 5-fold cross validation.
+For more details, see the vignette on the [glmnet website](https://glmnet.stanford.edu/articles/Coxnet.html).
+
+```{r glmnetCV}
+cvfit <- cv.glmnet(x_train, y_train, family = "cox", nfolds = 5)
+```
+
+## Model coefficients
+The coefficients (i.e., log hazard ratios) depend on the value of $\lambda$.
+We can easily view the coefficients for a specific value of $\lambda$ using `coef()`.
+We might want to examine coefficients from a model that minimizes the deviance:
+
+```{r coefLambdaMin}
+coef(cvfit, s = "lambda.min")
+```
+
+Or alternatively, we may choose a model that minimizes deviance within 1 standard deviation of the minimum (`lambda.1se`). Note that this results in a sparser model:
+
+```{r coefLambda1se}
+coef(cvfit, s = "lambda.1se")
+```
+
+The full regularization path can also be plotted:
+
+```{r plotCV}
+plot(cvfit)
+```
+
+# Model predictions and performance
+We will make out-of-sample predictions on the test set (25% of data) from models where the deviance was minimized (`lambda.min`).
+With survival data, there is no single best way to summarize predictions or to assess the performance of a model.
+We will first consider the C-index, which is a measure of the extent to which a model can *discriminate* across patients.
+Then we will examine survival probabilities and the extent to which predicted probabilities are well *calibrated*; that is, the extent to which predicted probabilities are  similar to observed probabilities.
+
+## C-index
+The C-index can be computed from the linear predictor (i.e., the predicted log hazards). By default, the `predict.cv.glmnet()` method predicts the linear predictor for each patient.
 
 ```{r}
-# Filter to patients alive at study entry --------------------------------------
-simdata_sub <- simdata %>% filter(hidden==0)
-
-# Split data -------------------------------------------------------------------
-# Using rsample package
-set.seed(7)
-data_split <- initial_split(simdata_sub, prop = .75)
-train_data <- training(data_split)
-test_data <- testing(data_split)
-
-# Clean up
-rm(data_split)
+lp_test <- predict(cvfit, newx = x_test, s = "lambda.min")
+head(lp_test)
 ```
 
-## Run Model
+The `survival::concordance()` function will compute the C-index in a manner appropriate for LTRC data.
+When making out of sample predictions, it is convenient to specify the `object` argument as a `formula` that relates the response to the linear predictor.
 
-Here we run a simple example penalized Lasso model for LTRC using `cv.glmnet()` in `glmnet` > 4.1.
-The lambda which minimizes the partial likelihood deviance (`lambda.min`) is chosen through 5-fold cross-validation.
-For more details, see the [glmnet homepage](https://glmnet.stanford.edu/articles/Coxnet.html).
+```{r cindex}
+concordance(y_test ~ lp_test, reverse = TRUE)
+```
+
+## Survival probabilities and calibration
+### Survival probabilities
+Models fit with `glmnet` now have `survfit` method which can be used to generate survival curves.
+Note that the values of `x` and `y` used to train the model must also be passed so that the baseline hazard can be computed.
 
 ```{r}
-# Prep x and y matrices - train
-x_train <- train_data %>% select(7:17) %>% as.matrix()
-y_train <- Surv(train_data$entry_time,
-          train_data$death_time,
-          train_data$dead)
-
-# Repeat for test data
-x_test <- test_data %>% select(7:17) %>% as.matrix()
-y_test <- Surv(test_data$entry_time,
-          test_data$death_time,
-          test_data$dead)
-
-# 5-fold CV glmnet
-mod <- cv.glmnet(x_train, y_train, family="cox", nfolds = 5)
-
+surv_test <- survfit(cvfit, newx = x_test, s = "lambda.min", 
+                    x = x_train, y = y_train)
 ```
 
-## Model Coefficients
+Plots using base `R` graphics can quickly be generating using the `plot.survfit()` method, which produces a survival curve for each patient in the test set.
 
-Choosing the best regularized model using `lambda.min`, we can assess model coefficients, which represent the log hazard ratios:
 ```{r}
-coef(mod, s = "lambda.min")
+plot(surv_test)
 ```
 
-Or alternatively, we may choose the model which minimizes deviance within 1 standard deviation of the minimum `lambda.1se`, which results in a sparser model:
-
-```{r}
-coef(mod, s = "lambda.1se")
-```
-
-The full path of regularization can also be plotted:
-```{r}
-plot(mod)
-```
-
-## Model Predictions and Performance
-
-We will make out-of-sample predictions on the 25% hold-out set using `lambda.min`. 
-```{r}
-predictions <- predict(mod, x_test, s="lambda.min")
-head(predictions)
-```
-
-### C-index
-
-```{r}
-get_cindex <- function(newx, newy, fit, lambda){
-  survival::concordance(newy ~ predict(fit, newx = newx, s = lambda), reverse=TRUE)
-}
-
-get_cindex(x_test, y_test, mod, "lambda.min")
-```
+Note that we could have alternatively used a general purpose plotting package like `ggplot` or specialized plotting package for survival analysis like `survminer` to produce more modern looking graphics.
+It might also be a good idea to aggregate predicted survival curves across patients in a meaningful way.
 
 ### Calibration
+While there are many great `R` packages such a `rms` and `pec` that can be used to evaluate a survival model (including for calibration), none (to our knowledge) work with LTRC data.
+We consequently wrote our own function to calibrate survival models for the paper (available in the file `R/calibrate.R`), which we will use here.
+`calibrate()` is a generic function with a method for `survfit` objects.
+We also created a generic `autplot()` method to quickly plot the output produced by `calibrate()`.
 
-```{r}
-# Predict survival probabilities
-survfit <- survfit(mod, newx = x_test, s="lambda.min", x= x_train, y= y_train)
+Let's calibrate the model at semi-annual intervals for up to 3 years.
+Each point consists of patients in a given decile of predicted survival at a particular time point.
+The x-axis is the average predicted survival probability and the y-axis is the pseudo observed survival probability computed using the (left truncation adjusted) Kaplan-Meier estimator.
+Perfectly calibrated predictions are those that lie along the 45-degree line so that predicted and observed survival probabilities are equal.
 
-# Get calibration at time points
-cal <- calibrate.survfit(object = survfit, times = seq(.5, 3, .5), y = y_test)
+```{r calibrate}
+# calibrate.survfit() method
+cal_test <- calibrate(object = surv_test, times = seq(.5, 3, .5), y = y_test)
 
-autoplot.calibrate(cal)
+# autoplot.calibrate() method
+autoplot(cal_test)
 ```

### 2021-06-18 simdata.R

@@ -1,0 +1,12 @@
+library("data.table")
+source("sim.R") # Contains functions used for the simulation
+
+# Simulate both predictors and survival times based on the CGDB
+sim_settings <- readRDS("sim_settings.rds")
+x_sim <- sim_x(n_pats = 1000, sim_settings, p_bin = 0)
+params <- set_params(x_sim, sim_settings, dist = "weibullPH")
+simdata <- sim_survdata(X = x_sim, params = params, n_pats = 10000)
+
+# Save
+simdata[, f_hidden := NULL]
+saveRDS(simdata, "simdata.rds")

### 2021-06-23 example.Rmd
Typo fixes for example

@@
 ### Survival probabilities
-Models fit with `glmnet` now have `survfit` method which can be used to generate survival curves.
+Models fit with `glmnet` now have a `survfit` method which can be used to generate survival curves.
 Note that the values of `x` and `y` used to train the model must also be passed so that the baseline hazard can be computed.
@@
 
-Plots using base `R` graphics can quickly be generating using the `plot.survfit()` method, which produces a survival curve for each patient in the test set.
+Plots using base `R` graphics can quickly be generated using the `plot.survfit()` method, which produces a survival curve for each patient in the test set.
 

### 2021-06-24 manuscript.tex
Update results with updated packages

@@
 The adjusted models tend to be better calibrated for patients with the lowest survival probabilities but poorly calibrated for patients with higher survival probabilities.
-One possible reason for this result is that the adjusted models pushed many of the hazard ratios to $1$ and did not adequately separate low and high risk patients.
+One possible reason for this result is that the adjusted models tended to push the most hazard ratios toward $1$, resulting in less separation of low and high risk patients.
 Probability rescaling such as Platt scaling or isotonic regression may be effective on the adjusted probabilities here, which resemble the characteristic sigmoid-shaped calibration curves produced by different learning algorithms, including the lasso; however the implementation is less straightforward in a survival context \cite{goldstein2021xcal}.
@@
 In the case of ``year of diagnosis'', this association was negative and biased the hazard ratio upwards.
-While many of the other hazard ratios in the large $p$ model were pushed towards 1 after left truncation adjustment, the direction of this bias depends on the direction of the association with left truncation time and varies by predictor.
+While some of the other hazard ratios in the large $p$ model were pushed towards 1 after left truncation adjustment, the direction of this bias depends on the direction of the association with left truncation time and varies by predictor.
 In cases where predictors and left truncation time are positively associated, hazard ratios are biased downwards because the predictors are associated with more immortal time.

### 2021-06-24 manuscript.tex
Minor edits to text

@@
 Penalized approaches, which add a regularization penalty to constrain the size of the $\beta$, reduce the complexity of the model and mitigate the negative effects of overfitting.
-Note that in high-dimensional settings where the number of predictors is larger than the sample size (i.e., $p$ > $n$), a standard regression cannot be fit at all.
+Note that in high-dimensional settings where the number of predictors is larger than the sample size (i.e.
+$p$ > $n$), a standard regression cannot be fit at all.
 
@@
 The observed survival time is given by $Y_i = \min(T_i, U_i)$ and a patient is right-censored if $T_i > U_i$.
-Patients are left-truncated (i.e., excluded from the sample) if $V_i > Y_i$.
+Patients are left-truncated (i.e.
+excluded from the sample) if $V_i > Y_i$.
 
@@
 
-Within the training set, Cox models with lasso penalties were fit among the ``observed'' (i.e., non-truncated) patients.
+Within the training set, Cox models with lasso penalties were fit among the ``observed'' (i.e.
+non-truncated) patients.
 Cross-validation with 10 folds was used to tune the  model and the value of $\lambda$ that minimized the partial likelihood deviance was selected.
@@
 
-The smaller model consisted of the following variables: year of diagnosis, age at diagnosis, race, practice type (academic or community center), and the mutational status (mutated/not mutated) of 3 known prognostic biomarkers for NSCLC (TP53, KRAS, and EGFR).
+The smaller model consisted of the following variables: year of diagnosis (i.e.
+index year), age at diagnosis, race, practice type (academic or community center), and the mutational status (mutated/not mutated) of 3 known prognostic biomarkers for NSCLC (TP53, KRAS, and EGFR).
 Short variants (SV) were considered for each biomarker while copy number alterations (CN) and rearrangements (RE) were also considered for TP53.
@@
 The arrows denote the impact of adjusting for left truncation on the hazard ratio.
-In some cases the adjustment can have a considerable impact on the hazard ratios, often moving a large hazard ratio all the way to $1$ (i.e., to a coefficient of $0$).
+In some cases the adjustment can have a considerable impact on the hazard ratios, often moving a large hazard ratio all the way to $1$ (i.e.
+to a coefficient of $0$).
 This is most notable with diagnosis year, which has a hazard ratio of approximately $1.5$ in both the small and large $p$ models but is reduced to $1$ or less than $1$ after the left truncation adjustment.
@@
 The adjusted models tend to be better calibrated for patients with the lowest survival probabilities but poorly calibrated for patients with higher survival probabilities.
-One possible reason for this result is that the adjusted models tended to push the most hazard ratios toward $1$, resulting in less separation of low and high risk patients.
+One possible reason for this result is that the adjusted models tended to push hazard ratios---especially for the influential ``year of diagnosis'' predictor---toward $1$, resulting in less separation of low and high risk patients.
 Probability rescaling such as Platt scaling or isotonic regression may be effective on the adjusted probabilities here, which resemble the characteristic sigmoid-shaped calibration curves produced by different learning algorithms, including the lasso; however the implementation is less straightforward in a survival context \cite{goldstein2021xcal}.
@@
 On the other hand, metrics that focus on predicted survival probabilities such as calibration curves are better able to determine whether a model can accurately predict the risk of an event.
-
 In our real-world use case, an analyst could have mistakenly concluded that the C-index in the non-adjusted model was indicative of good model performance when it was in fact driven in part by a high correlation between predictors and left truncation.
@@
 When making predictions with survival data is of interest, it is critical to define the risk set appropriately, as our work demonstrates.
-To do this requires recognizing the underlying process by which individuals are captured in the data in the first place; i.e., when and how individuals came to be included in the dataset of interest.
+To do this requires recognizing the underlying process by which individuals are captured in the data in the first place; i.e.
+when and how individuals came to be included in the dataset of interest.
 One of the motivations for the current work derives from the advent of biobank and genomic test databases that store information on patient illness - presumably after the patient has been diagnosed and ``at risk'' for some time.

### 2021-06-25 manuscript.tex
tweak description of year HR

@@
 to a coefficient of $0$).
-This is most notable with diagnosis year, which has a hazard ratio of approximately $1.5$ in both the small and large $p$ models but is reduced to $1$ or less than $1$ after the left truncation adjustment.
+This is most notable with diagnosis year, which has a hazard ratio of approximately $1.5$ in both the small and large $p$ models but is reduced to less than $1$ after the left truncation adjustment.
 This effect is likely due to the strong negative correlation between calendar time and the probability that a patient is left-truncated.
@@
 In our real-world use case, an analyst could have mistakenly concluded that the C-index in the non-adjusted model was indicative of good model performance when it was in fact driven in part by a high correlation between predictors and left truncation.
-For instance, in our large $p$ model the hazard ratio for the predictor ``year of diagnosis'' moved from being the largest in absolute value to around 1 after adjusting for left truncation.
+For instance, in our large $p$ model the hazard ratio for the predictor ``year of diagnosis'' moved from being the largest in absolute value to less than 1 after adjusting for left truncation.
 The strong positive relationship between ``year of diagnosis'' on the hazard in the non-adjusted model was artificial because patients diagnosed earlier appeared to survive longer due to immortal time bias.

### 2021-07-20 example.Rmd
Minor edits to prose in example

@@
 ### Input matrix
-`glmnet` requires an input matrix `x`, which which must be a `matrix` object. Here we consider a simple example with a small number of predictors contained in `x`.
+`glmnet` requires an input matrix `x`, which must be a `matrix` object. Here we consider a simple example with a small number of predictors contained in `x`.
 
@@
 When fitting Cox proportional hazards models with `glmnet`, the response should be a `survival::Surv` object.
-The key difference between modeling LTRC and RC data, is that the response must contain start-stop intervals when using LTRC data.
-Start-stop `Surv` object are created by passing the time that a patient enters the dataset (`time`), the follow-up time (`time2`), and a status indicator equal to 1 if a patient has died and 0 if they are alive.
+The key difference between modeling LTRC and RC data is that the response must contain start-stop intervals when using LTRC data.
+Start-stop `Surv` objects are created by passing the time that a patient enters the dataset (`time`), the follow-up time (`time2`), and a status indicator equal to 1 if a patient has died and 0 if they are alive.
 
@@
 ## Cross validation
-We will fit a penalized Lasso model appropriate for LTRC data using `cv.glmnet()` (note: requires `glmnet` > 4.1).
-The value of the tuning parameter$\lambda$ that minimizes the partial likelihood deviance (`lambda.min`) is selected using 5-fold cross validation.
+We will fit a penalized lasso model appropriate for LTRC data using `cv.glmnet()` (note: requires `glmnet` > 4.1).
+The value of the tuning parameter $\lambda$ that minimizes the partial likelihood deviance (`lambda.min`) is selected using 5-fold cross validation.
 For more details, see the vignette on the [glmnet website](https://glmnet.stanford.edu/articles/Coxnet.html).
@@
 
-Plots using base `R` graphics can quickly be generated using the `plot.survfit()` method, which produces a survival curve for each patient in the test set.
+Plots using base `R` graphics can be quickly generated using the `plot.survfit()` method, which produces a survival curve for each patient in the test set.
 
@@
 
-Note that we could have alternatively used a general purpose plotting package like `ggplot` or specialized plotting package for survival analysis like `survminer` to produce more modern looking graphics.
+Note that we could have alternatively used a general purpose plotting package like `ggplot` or a specialized plotting package for survival analysis like `survminer` to produce more modern looking graphics.
 It might also be a good idea to aggregate predicted survival curves across patients in a meaningful way.
@@
 ### Calibration
-While there are many great `R` packages such a `rms` and `pec` that can be used to evaluate a survival model (including for calibration), none (to our knowledge) work with LTRC data.
+While there are many great `R` packages such a `rms` and `pec` that can be used to evaluate a survival model (including via calibration), none (to our knowledge) work with LTRC data.
 We consequently wrote our own function to calibrate survival models for the paper (available in the file `R/calibrate.R`), which we will use here.

### 2021-07-28 example.Rmd
Add link to paper in example.Rmd

@@
 # Introduction
-This guide shows how to use penalized survival models for prediction with left-truncated and right-censored data (LTRC) as described in McGough *et al.* 2021.
+This guide shows how to use penalized survival models for prediction with left-truncated and right-censored data (LTRC) as described in [McGough *et al.* 2021](https://onlinelibrary.wiley.com/doi/10.1002/sim.9136).
 The analyses utilize a simulated dataset based on the Foundation Medicine and Flatiron Health Clinico-Genomic Database (CGDB), as described in the paper.
